\documentclass[11pt, oneside]{article}   	% use ``amsart'' instead of ``article'' for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --\rangle pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{verbatim}
\usepackage{tikz} 

\usepackage{syntonly}
% \syntaxonly \langle -- use this for checking syntax only
% \mbox {text} - keep together
% \fbox {text} - keep together and draw around

%\pagestyle{plain|headings|empty} % header and footer p.27
%SetFonts
%\include{filename}, \includeonly{filename1, filename2} , \input[fiename}

%SetFonts% 

\title{Brilliant: Vector Calculus}
\author{Dave Fetterman}
\date{6/21/22}							% Activate to display a given date or no date

\begin{document}
\maketitle
Note: Latex reference: http://tug.ctan.org/info/undergradmath/undergradmath.pdf

\section{Chapter 2: Basics, Vector fields}
\subsection{Chapter 2.1: Calculus of Motion}

Consider vectors of motion against $t$ of the form $ \vec{x}(t) = \langle x(t), y(t), \ldots \rangle$.
\begin{itemize}
\item A \textbf{line} through $p = (a, b, c)$ parallel to $\vec{v} = \langle v_x, v_y, v_z\rangle$ is $\vec{x}(t) = \vec{p} + t \vec{v}$ 
\item \textbf{velocity} is characterized completely by $\vec{v}(t) = \vec{x}'(t) = \langle x'(t), y'(t), z'(t)\rangle$.
\item The \textbf{speed} of an object along that line versus $t$ is the length of $v$ ($\|v\|$)  
\item Therefore, the speed of an object along line 
$$ \langle x(t), y(t), z(t)\rangle = \langle 0, 2, -3\rangle + t\langle 1,-2,2\rangle$$ is $$\sqrt{1^2+(-2)^2 + 2^2} = 3$$
\item Note that $\vec{v}$ need not be constant.  The speed of $$\vec{x}(t) =  \vec{p} + 3 \sin(2\pi t)\hat{u}, \| \hat{u} \| = 1$$ would then be $$\| 6\pi \cos(2 \pi t) \hat{u} \| = |6\pi \cos(2 \pi t)|$$
\item \textbf{Acceleration} $a(t) = v'(t) = x''(t)$ is straightforward.  Acceleration of $$x(t) = \langle -1 + \cos(t), 1, \cos(t)\rangle = \langle -\cos(t), 0, -\cos(t)\rangle$$
\item An example position vector for a planet of distance $r$ from the sun could be $\langle r \cos(t), r \sin(t) \rangle$.  The acceleration vector points in the opposite direction: $\langle - r \cos(t), - r \sin(t) \rangle$.  In addition to being the analytical second derivative, consider that the \emph{force} of gravity, (which, by $F = ma$ is proportional to acceleration) points towards the sun, \emph{with acceleration perpendicular to velocity}.

\item A \textbf{helix} could be a 3D extension like $\langle r \cos(t), r \sin(t), b\cdot t \rangle$.  
\end{itemize}

\section{Chapter 2.2: Space Curves}
\begin{itemize}
\item Note that while $\vec{x}(t) = \langle \cos(t), \sin(t), 5 \rangle$ and  $\vec{x}(t) = \langle \cos(2t), \sin(2t), 5 \rangle$ describe the same curve, the space curve also records motion in time, so the \emph{velocity} may be different.
\item If $\vec{x}(t) = t\vec{v}$, then speed is $\frac{\| \vec{x}(t+\Delta t) - \vec{t} \|}{\Delta t} = \| \vec{v} \|$,  direction is $\frac{\vec{v}}{\| \vec{v} \|}$, and velocity $\vec{v}$ is the product of speed and direction.
\item So $\vec{v}(t) = \lim_{\Delta t \rightarrow 0} \frac{\vec{x}(t + \Delta t) - \vec{x}(t)}{\Delta t} = \vec{x}'(t) = \frac{d\vec{x}}{dt} = \langle x'(t), y'(t), z'(t) \rangle$
\item Neat conceptual result: any $y = f(x)$ can be made into $x(t) = \langle t, f(t) \rangle$, and then $v(t) = \langle 1, f'(t) \rangle$, which points along the tangent line at $\langle t, f(t)\rangle$.
\item Note that dot product derivatives work like regular product: $[\vec{a}(t) \cdot \vec{b}(t)]' = \vec{a}'(t) \cdot \vec{b}(t) + \vec{a}(t) \cdot \vec{b}'(t)$, 
but the cross product does not work the same since $\frac{d}{dt}[a \times b] = a' \times b + a \times b'$, but since $a \times b' = -b' \times a$, can't switch the order to $a' \times b + b' \times a$ due to this non-commutativity.
\item If $$\vec{x}(t) = \vec{p}+ t \vec{v},$$ calculating velocity with respect to origin  becomes
$$ \frac{d}{dt} \| \vec{x}(t) \| = \frac{\vec{x}(t) \cdot \vec{x}'(t)}{\| \vec{x}(t) \|} = \frac{\vec{x}}{\| \vec{x} \|} \cdot \vec{v},$$ after rewriting the distance formula and chugging through the chain rule.
\item However, it becomes more clear when considering that $(\vec{v} \cdot \hat{x}) \hat{x}$ is the projection of the velocity vector onto the position vector.  So, the length of this is the rate of change of distance from origin!
\end{itemize}

\section{Chapter 2.3: Integrals and Arc Length}
\begin{itemize}
\item Integral of a vector function can be defined componentwise in a straightforward way: $\int_a^b {\vec{x}(t)} = \langle \int_a^b {x(t)},  \int_a^b {y(t)},  \int_a^b {z(t)} \rangle $ 
\item Example: if ball launched from origin with velocity $\langle 1, 2, 3 \rangle$ and acceleration $\langle 0, 0, -1 \rangle$, it lands at
\begin{align} 
\frac{dv}{dt}dt = \langle 0, 0, -1 \rangle \\ 
\int{\frac{dv}{dt}dt} = v =  \langle C , D , -t +F \rangle  = \langle 1, 2, 3 \rangle =  \langle 1, 2, -t + 3\rangle, t = 0 \\
x = \int{v}  = \langle t+K, 2t+M, -\frac{1}{2}t^2 + 3t + N \rangle, x(\vec{0}) = \langle 0 ,0, 0\rangle \\ 
\vec{x}(t) = \langle t, 2t, 3t-\frac{1}{2}t^2 \rangle \\
z(t) = 0 \rightarrow t = 6 \rightarrow \vec{x}(6) = \langle 6, 12, 0 \rangle \\
\end{align} 
\item Also, generalizing $ds = \sqrt{(dx)^2  + (dy)^2}$, the length of an arc from point $a$ to $b$ approaches \fbox{$\int_a^b ds = \int_{t_a}^{t_b} \| x'(t) \| dt$}
\item Example: a helix $\langle a \cos (\omega t), a \sin (\omega t), b \omega t \rangle$, parametrized by time $t$ can be rewritten in terms of $s$, the arc length: 
\begin{align}
s =  \int \|x'(t)\| dt \\
s = \int \sqrt{(-\omega a \sin (\omega t))^2 + (\omega a \cos (\omega t))^2 + (b \omega)^2}dt \\
s = |\omega| \int \sqrt{(a^2 + b^2)}dt \\ 
s = |\omega| t \sqrt{a^2 + b^2}
\end{align}
\item \emph{Note: It's weird to think of time in terms of length}.  Could be analytically useful?
\end{itemize}


\section{Chapter 2.4: Frenet Formulae}

Main idea: Establish three new vectors $\hat{T}(s), \hat{N}(s), \hat{B}(s)$ that change as we move along a space curve, instead of $\vec{x}(t)$ that changes over an external ``time'' idea.

Remember that  $s = \int_0^t \| \vec{x}'(\tilde{t}) \| d\tilde{t}$, so $\frac{ds}{dt} = \| \vec{x}'(t)\|$ .

\subsection{$\hat{T}$: Vector tangent to space curve}
\begin{itemize}
\item Remember arc length is $s = \int_0^t \| \vec{x}'(\tilde{t})d\tilde{t} \|$
\item $\hat{T}$ is just normalized grad: $\frac{\vec{x}'(t)}{\|\vec{x}'(t)\|}$
\item This implies \fbox{$\frac{d\vec{x}}{ds} = \hat{T}$} since 
\begin{align}
s = \int_0^t \| \vec{x}'(\tilde{t})d\tilde{t} \| \\
\frac{ds}{dt} = \| \vec{x}(t) \| \\
\hat{T} = \frac{\vec{x}'(t)}{\|\vec{x}'(t)\|}  = \frac{d\vec{x}}{dt} \cdot \frac{dt}{ds} \\
\hat{T} = \frac{d\vec{x}}{ds} \\
\end{align}
\item So this is how the space curve $\vec{x}$ changes as it moves along the curve at length $s$.
\item It's normalized, so it's the same whether parameterized by t, s, or whatever.
\end{itemize}


\subsection{$\hat{N}$: Normal to curve (perpendicular to $\hat{T}$)}
Normal vectors are:
\begin{itemize}
\item $\vec{x}''(t)$ normalized as \fbox{$\frac { \frac{d\hat{T}}{ds} }   {\|  \frac{d\hat{T}}{ds}  \|}  = \hat{N}$}
\item The normal vector to the curve
\item $\bot$ to $\hat{T}$ in direction of acceleration.  So a multiple of acceleration vector.
\item $\frac{\hat{T}'(t)} { \| \hat{T}'(t) \|}$.  The following sequence shows any unit length vector is perpendicular to its derivative.
\begin{align}
\| \hat{T} \| = 1 \\
d(\| \hat{T} \|^2) = 0 \\
d(\| \hat{T} \|^2) = d(\hat{T} \cdot \hat{T}) = \hat{T}(t) \cdot 2\hat{T}'(t) \\
\hat{T}(t) \cdot \hat{T}'(t)  = 0
\end{align}
\item $\frac{ \frac{d\hat{T}}{ds}  }  { \| \frac{d\hat{T}}{ds}  \| }$ since it's the same as the above, but parametrized over $s$ instead of $t$.  Doesn't change the direction of the vector!
\end{itemize}

Example:if  $\vec{x}(t) = \langle R\cos(\omega t), R\sin(\omega t), 0 \rangle$, then:
\begin{itemize}
\item $\vec{a} = \frac{d^2\vec{x}}{dt^2}$ just by definition
\item $\vec{a} = -\omega^2 \vec{x}$ just by calculation
\item $\hat{T}(t) = \langle -\sin(\omega t), \cos (\omega t), 0 \rangle$
\item $\| \hat{T}(t) \| = 1$
\item $\hat{N} = \frac{\hat{T}'(t)} { \| \hat{T}'(t) \|} =  \langle -\cos(\omega t), -\sin (\omega t), 0 \rangle$
\item So $\vec{a} = R\omega^2\hat{N}$ by these formulae.
\end{itemize}

This leads us to believe acceleration and $\hat{N}$, the normed derivative of $\hat{T}$ are related.

The part of acceleration $\vec{a}$ parallel to $\hat{T}$ is the projection $(\vec{a} \cdot \hat{T}) \hat{T}$

The perpendicular part is then $\vec{a}$ minus that: $\vec{a} - (\vec{a} \cdot \hat{T}) \hat{T}$

This also equals $(\frac{ds}{dt})^2 \|\frac{d\hat{T}}{ds} \| \hat{N}$ because
\begin{align}
\vec{x}' = \frac{dx}{dt} = T = \hat{T} \cdot \|  \frac{dx}{dt} \| \\
s = \int_0^t \| \vec{x}'(t) \| \rightarrow \frac{ds}{dt} = \| \vec{x}'(t) \|  
\end{align}

$\hat{N} = \frac{d\hat{T}}{ds}$ normalized, so
\begin{align}
\vec{a} = \frac{d^2\vec{x}}{dt^2} = \frac{d}{dt}(\|\frac{dx}{dt}\|  \frac{\frac{dx}{dt}} {\|\frac{dx}{dt}\|} ) = \frac{d}{dt}(\| \vec{x}'(t)\|\hat{T}(t)) = \frac{d \|\vec{x}'(t) \|}{dt}\hat{T} + \| \vec{x}'(t) \| \frac{d\hat{T}}{dt} \\
= \frac{d \|\vec{x}'(t) \|}{dt}\hat{T} + \frac{ds}{dt} \frac{d\hat{T}}{ds} \frac{ds}{dt} \\ 
= \frac{d \|\vec{x}'(t) \|}{dt}\hat{T} + (\frac{d{s}}{dt} )^2 \| \frac{d\hat{T}}{ds} \| \hat{N}
\end{align}

This is ``a = $\hat{T}$'s parallel part plus $\hat{T}$'s perpendicular (N) part'', so the second term is $a_{\bot}$

\subsection{$\hat{T}$ and $\hat{N}$}

\begin{itemize}
\item Form a plane, since first, any normal vector's derivative is perpendicular to the vector
\item $\kappa$ is curvature: how much we're curving in that $T \times N$ plane.
\item \fbox{$\kappa =   \|  \frac{d\hat{T}}{ds}  \| $ }
\item Therefore, by the definition of $\hat{N} = \frac{d\hat{T}/ds}{\|d\hat{T}/ds\|}$, \fbox{$\frac{d\hat{T}}{ds} = \kappa \hat{N}$ }\textbf{(Frenet equation 1)}
\end{itemize}

Note that curvature $\kappa(s) = \| \frac{d\hat{T}}{ds}\|$ is geometric (depends on s, not time) and changes as $\hat{T}$ changes.

Example: Curvature of $\vec{x}(t) = \langle \cos(t), \sin(t), bt \rangle$ 
\begin{align}
x'(t) = \langle -\sin(t), \cos(t), b \rangle \\ 
\|x'(t) \| = \sqrt{1 + b^2} \\ 
s = \int_0^t\|x'(\tilde{t})\| d\tilde{t} = \int_0^t \sqrt{(1+b^2)} = t\sqrt{(1+b^2)} \rightarrow t = \frac{s}{\sqrt{1+b^2}}
\end{align}

Do the substitution of $\frac{s}{\sqrt{1+b^2}}$ for $t$ above to get $x'(s)$, and from there, you can figure out $\frac{d\hat{T}}{ds}$ and normalize to get $\kappa = \frac{1}{1+b^2}$

\subsection{$\hat{B}$ is binormal: perpendicular to both}
\begin{itemize}
\item defined as \fbox{$\hat{B} = \hat{T} \times \hat{N}$}
\item Therefore, by derivative 
\begin{align}
\frac{d\hat{B}}{ds} = \frac{d\hat{T}}{ds} \times \hat{N} +  \hat{T} \times \frac{d\hat{N}}{ds}\\ 
\frac{d\hat{B}}{ds} = \kappa \hat{N} \times \hat{N}  +  \hat{T} \times \frac{d\hat{N}}{ds}\\ 
\frac{d\hat{B}}{ds} =  \hat{T} \times \frac{d\hat{N}}{ds}
\end{align}
This means T is orthogonal to dB, and we already know B and dB are orthogonal.
We're working in 3D with the cross product, so dB is parallel to N.
\item Therefore, we define \textbf{torsion} $\tau$ so that \fbox{$- \frac{d\hat{B}}{ds}  = \tau \hat{N}$  }\textbf{(Frenet equation 2)}.  Negative sign by convention.
\item Can also cross by $N$ on both sides to get $- \frac{d\hat{B}}{ds}  \times \hat{N} = \tau $ 
\item $\tau$ measures how the plane defined by $\hat{T}, \hat{N}$ twists around.  On a circle, $\hat{B}$ wouldn't change, so the derivative would be zero.
\item \textbf{Final Frenet equation}.  Prereq: $\hat{B} = \hat{T} \times \hat{N} \rightarrow \hat{N} = \hat{B} \times \hat{T} \rightarrow \hat{T} = \hat{N} \times \hat{B}$
\begin{align}
\frac{d\hat{N}}{ds} = \frac{d\hat{B}}{ds} \times \hat{T} +   \hat{B} \times  \frac{d\hat{T}}{ds} \\
\frac{d\hat{N}}{ds} = -\tau \hat{N} \times \hat{T} + \hat{B} \times \kappa \hat{N} \\ 
\frac{d\hat{N}}{ds}  = \tau \hat{B} - \kappa \hat{T}
\end{align}

\end{itemize}

\section{Chapter 2.5: Parametrized Surfaces}

Main approaches to describing a surface:
\begin{itemize}
\item Can parameterize by $\vec{x}(u,v) = x(u,v), y(u,v), z(u,v)$
\item Can perhaps parameterize $f(x,y,z) = c$ by $z = g(x,y)$
\item Can also use ideas like $\nabla f = 0$ to find a normal.
\end{itemize}

There are many out-of-the-box paremetrizations including:

\begin{itemize}
\item Sphere at (0,0,0): $\vec{x}(u,v) = \langle R\cos(u)\sin(v), Rsin(u)\sin(v), R\cos(v)\rangle$, where $u \in [0, 2\pi), v \in [0, \pi]$
\item Rotate function $y = f(x)$ around the x-axis: $\vec{x}(u,v) = \langle u, f(u)\cos(v), f(u)\sin(v) \rangle$, where $u \in D, v \in [0, 2\pi]$
\end{itemize}

Tangent vectors to $\vec{x}(u,v)$ are $\frac{\delta \vec{x}} {du}, \frac{\delta \vec{x}} {dv}$, so unit normal \fbox{	$\hat{n} = \pm \frac 
					{\frac{d \vec{x}} {du} \times \frac{\delta \vec{x}}{dv}}
					{\| \frac{d \vec{x}} {du} \times \frac{\delta \vec{x}}{dv} \|}
					$}
					
\textbf{Example}: Torus $\vec{x} = \langle [2 + \cos(v)]\cos(u),  [2 + \cos(v)]\sin(u), \sin(v)\rangle, u,v \in [0,2 \pi)$.  What's the tangent plane at $u = \frac{\pi}{4}, v=0$?
\begin{align}
d\vec{x}/du = \langle -\sin(u)(2 + \cos(v)),  \cos(u)(2 + \cos(v)), 0 \rangle \\
d\vec{x}/dv = \langle -\sin(v)\cos(u), -\sin(v)\sin(u), \cos(v) \rangle \\
u = \frac{\pi}{4}, v=0 :  \\
d\vec{x}/du = \langle -\frac{3}{\sqrt{2}}, \frac{3}{\sqrt{2}}, 0, \rangle \\
d\vec{x}/dv = \langle 0, 0, 1 \rangle \\
dx/du \times dx/dv = \langle  \frac{3}{\sqrt{2}}, \frac{3}{\sqrt{2}}, 0\rangle \\
\hat{n} =  \langle \frac{1}{\sqrt{2}}, \frac{1}{\sqrt{2}}, 0 \rangle \\
\hat{n}  \cdot \vec{x} = 0 \rightarrow \hat{n} \cdot (x - x_0, y-y_0, z-z_0) = 0 \\
\rightarrow ... \rightarrow x + y = 3\sqrt{2}  \\
\end{align}



\subsection{Example: Ellipsoid $x^2 + 2y^2 + z^2 = 4$ What's the normal at $(1, \frac{1}{\sqrt{2}}, \sqrt{2})$?}

\textbf{Method 1: parametrize with spherical u, v}
First, transform to sphere with change of coordinates, then flip to speherical coordinates.
\begin{align}
x^2 + 2y^2 + z^2 = 4 \\
X = x/2, Y=\frac{Y}{\sqrt{2}}, Z = z/2 \\
X^2 + Y^2 + Z^2 = 1 \\
X = \cos(u)\sin(v), Y = \sin(u)\sin(v), Z = \cos(v) \\
p = (1, \frac{1}{\sqrt{2}}, \sqrt{2})  \rightarrow u = v = \frac{\pi}{4} \\
\frac{dx}{du}(\frac{\pi}{4}, \frac{\pi}{4}) = \langle -1, \frac{1}{\sqrt{2}}, 0 \rangle \\
\frac{dx}{dv}(\frac{\pi}{4}, \frac{\pi}{4}) = \langle 1, \frac{1}{\sqrt{2}}, -\sqrt{2} \rangle \\
\frac{dx}{du}(\frac{\pi}{4}, \frac{\pi}{4}) \times \frac{dx}{dv}(\frac{\pi}{4}, \frac{\pi}{4}) = \langle 1, \sqrt{2}, \sqrt{2}\rangle \\
\hat{n}_{out} = \frac{\langle -1, -\sqrt{2}, -\sqrt{2}\rangle }{\sqrt{5}}
\end{align}


\textbf{Method 2: rewrite as z = g(x,y)}

\begin{align}
x^2 + 2y^2 + z^2 = 4 \\
z = (4 - x^2 - 2y^2)^\frac{1}{2} \\
dz / dx = \frac{1}{2} \times -2x (4 - x^2 - 2y^2)^{-\frac{1}{2}} = -\frac{1}{\sqrt{2}} \\
dz / dy = \frac{1}{2} \times -4y (4 - x^2 - 2y^2)^{-\frac{1}{2}} =-2\sqrt{2}/\sqrt{2} = -1\\
f \approx \sqrt{2}  + dz / dx(1, \frac{1}{\sqrt{2}})(x - 1) + dz / dy (1,  \frac{1}{\sqrt{2}}) (y - \frac{1}{\sqrt{2}})\\
\rightarrow ... \rightarrow \frac{1}{\sqrt{2}}x + y + z = 2\sqrt{2} \\
\end{align}

giving us normal vector $\langle \frac{1}{\sqrt{2}}, 1, 1 \rangle = \frac{\langle 1, \sqrt{2}, \sqrt{2}\rangle }{\sqrt{5}}$
after normalization.

\textbf{Method 3: gradient}

Gradient is always normal to the tangent plane.
Recognize level set of $f(x,y,z) = x^2 + 2y^2 + z^2$.  

$\nabla f = \langle 2x, 4y, 2z \rangle \rightarrow \nabla f(1, \frac{1}{\sqrt{2}}, \sqrt{2}) = \langle 2, 2\sqrt{2}, 2\sqrt{2}\rangle$

Then normalize.

\subsection{Mobius strip and ``outward direction''}
Mobius strip is 
\begin{itemize}
\item $x = 2\cos(u) + v\cos(\frac{u}{2})$
\item $y = 2\sin(u) + v\cos(\frac{u}{2})$
\item $z = v \sin(\frac{u}{2})$
\item $u \in [0, 2\pi], v \in [-\frac{1}{2}, \frac{1}{2}]$
\end{itemize}

$\hat{n} = \frac
	{\vec{x}_u \times  \vec{x}_v}
	{\|\vec{x}_u \times  \vec{x}_v\|}
	$ at (0,0) is $\langle 0, 0, -1 \rangle$, 
	
but at the same point $(2\pi, 0)$ $\hat{n} = \langle 0, 0, 1 \rangle$!!
	


\section{Chapter 2.6: Vector Fields}

(Lots of intuition questions here...)

One nugget: using \textbf{gradient vector fields}:
Suppose $\vec{F}(x,y) = \langle 2, -4y^3 \rangle$.  
If $\vec{F} = \nabla f$ for some (single value function) $f$, then $F$'s arrows are perpendicular to a level set $f = c$.
So look at $f  =  2x -y^4 $ and find perpendicular arrows to these.  That's actually F!

\textbf{Linear approximation for $\vec{F}: D \in \mathbf{R}^n \rightarrow \mathbf{R}^m$}

Main idea: 
$\vec{F}(\vec{x}) = \vec{F}(\vec{a}) + A(\vec{a})(\vec{x} - \vec{a})$

Note that $A$ takes in vectors of size $n$ (so it has as many columns as the input space), and has $m$ functions (rows) that operate on it.
So the Jacobian, $A$, has as row $i$, column $j$, the quantity $\frac{dF_i}{dx_j}(\vec{a})$.

$dF_i/d\vec{x}$ extends across row $i$.

\section{Chapter 2.7: Jack and the Beanstalk (Newton's method)}

\textbf{Basis for Newton's:}

If we're estimating $x_1$ by following the derivative at $x_0$, this means we're looking at the line with x-intercept $x_1$, with slope $f'(x_0)$.

  So instead of $y = mx+b$, we'll flip the two and use 
  
 $x = y/m + x_{int}$ 
 
or $x_0 = f(x_0)\frac{1}{f'(x_0)} + x_1$, 

or \fbox{$x_1 = x_0 - \frac{f(x_0)}{f'(x_0)}$}

Note that, under Newton's something like $|x|$ will converge immediately, $x^3$ will converge moderately, and a S-curve might barely converge if at all.

The extension of this with the Jacobian matrix $A = DF'(x_0)$ is $\vec{x}_1 = \vec{x}_0 - (D\vec{F}(\vec{x}_0))^{-1}\vec{F}(x_0)$ 


\section{Chapter 2.8: Electrostatic bootcamp}

Electric charge radiates out equally in all directions, and is inversely proportional to distance.  

Formula, with $Q$ as the charge, $\epsilon_0$ is a constant: $\vec{E}(x,y,z) = \frac{Q}{4\pi\epsilon_0\|x\|^2}\hat{x}$

A field line is a special case of a \textbf{flow line} - the space curve that follows $\vec{F}$'s arrows.  The tangent vector to the flow line is $\vec{F}(\vec{x}(\tilde{t}))$ ($\tilde{t}$ is not time here), so $\frac{d\vec{x}}{d\tilde{t}} = \vec{F}(\vec{x}(\tilde{t}))$

Example: Vector field $\vec{F}(x,y) = \langle -2y, 3x \rangle$.  What's the flow line through (2,0)?

Solution: Need to solve 
$dx/dt = -2y, dy/dt = 3x$.  Key is ``separating the equations''.  Remember x and y are functions of t!

\begin{align}
\frac{d^2x}{dt^2} = -2 \frac{dy}{dt}= -2 \times 3x = -6x.\\
\frac{d^2y}{dt^2} = -2 \frac{dy}{dt}= -2 \times 3x = -6y.\\
x(t) = -6x''(t), y(t) = -6y''(t) \\
\rightarrow x = A\cos(\sqrt{6}t)+ B\sin(\sqrt{6}t),  y = C\cos(\sqrt{6}t)+ D\sin(\sqrt{6}t) \\
\frac{dx}{dt} = -2y(t) \rightarrow   \frac{\sqrt{6}}{2}A\sin(\sqrt{6}t)- \frac{\sqrt{6}}{2}B\cos(\sqrt{6}t)  = y(t) \\
x(t=0) = 2 \rightarrow A = 2 \\
y(t=0) = 0 \rightarrow B = 0 \\
\vec{F}(t) = \langle 2\cos(\sqrt{6}t), \sqrt{6}\sin(\sqrt{6}t) \rangle \\
\end{align}

Note: \textbf{Field lines} follow rules:
\begin{itemize}
\item Go from positive charges to negative
\item Density of lines directly relates to how much charge a point has
\item Lines don't intersect.
\item Corollary: If count of out equals count of in, point has zero charge
\item ``Number'' (to be defined) of field lines in and out of a \emph{surface} related to the charge inside.  Upcoming.
\end{itemize}

\section{Chapter 3: Surface integrals, Flux, Divergence}

\subsection{3.1: Surface Integrals}

Example: Fluid pressure in a tank is:
\begin{itemize}
\item Proportional (via some weight constant $p_{fluid}$) to depth of the point
\item Pushes out via the normal $\hat{n}$ 
\item So, for the $x=l$ side of a cube of length l, this would be 

$\vec{F}_{x=l} = (\iint_{[0,l] \times [0,l]} p_{fluid} [1 - \frac{z}{l}] dy dz) \hat{i}$
\end{itemize}



\textbf{Example}: Hemisphere of size l, sitting at (0, 0, 0)

Finding the out pointing unit normal of hemisphere at point $(x, y, \sqrt{l^2-x^2-y^2})$

Note: Can just eyeball this, but one way is the \textbf{gradient}.  

First, the relation is $x^2 + y^2 + (z-l)^2 = l^2$.
Make it a function $g$ and take the level set at $l^2$:

\begin{align}
g(x,y,z) = x^2  + y^2 + (z - l)^2 = l^2 \\
\nabla g(x,y,z) = \langle 2x , 2y , 2(z-l)  \rangle \\
\hat{n} = \pm \frac{\nabla g(x,y,z)}{ \|\nabla g(x,y,z) \|} \\ 
\hat{n} = \pm \frac{\langle x , y , (z-l)  \rangle}{\sqrt{x^2 + y^2 + (z-l)^2}} \\
\hat{n} = \pm \frac{\langle x , y , (z-l) \rangle}{\sqrt{l^2}} \\
\hat{n} = \pm \langle \frac{x}{l}, \frac{y}{l} ,\frac{z}{l} - 1\rangle \\
\end{align}


Note: Integrating over a patch dA on the surface means finding the area of micro-patches 
$\Delta A_{ij}$, which is the parallelogram defined by 
\begin{align}
s_1 = \langle \Delta x_i, 0, \Delta x_i f_x(x_i^*,y_j^*) \rangle \\
s_2 = \langle 0, \Delta y_j, \Delta y_j f_y(x_i^*,y_j^*) \rangle \\ 
\Delta A_{ij} \approx \| s_1 \times s_2 \| \\ 
= \sqrt{ (1 + [ f_x(x_i^*,y_j^*)]^2 + [f_y(x_i^*,y_j^*)]^2 }\Delta x_i \Delta y_j \\ 
\end{align}

So if \fbox{$z = f(x,y), dA = \sqrt{1 + f_x^2 + f_y^2}$}.


\textbf{So the total pressure ends up being}
$\vec{F}_{tot} = p_{fluid} \iint (p \cdot \hat{n}) dA$
\begin{align}
= p_{fluid} \iint_{x^2 + y^2 \leq l^2} [ 1- \frac{f(x,y)}{l}] \hat{n}  \sqrt{ 1 +  [f_x]^2 +[f_y]^2 }dxdy  \\
f(x,y) = l - \sqrt{l^2-x^2-y^2} \\
\hat{n} = \langle \frac{x}{l},\frac{t}{l}, \frac{f(x,y)}{l} - 1 \rangle \\
f_x = \frac{x}{\sqrt{l^2 - x^2 - y^2}}, f_y = \frac{y}{\sqrt{l^2 - x^2 - y^2}}
\end{align}

And for the only-nonzero component, $\hat{k}$, this simplifies after a lot of hand-math to 
$F_{tot} = -p_{fluid} ( \iint_{x^2+y^2 \leq l^2} \sqrt{1 - (\frac{x^2 + y^2}{l^2})} dx dy) \hat{k}$

Side Note during solving: $dx dy \rightarrow r dr d\theta$.  
\begin{itemize}
\item TODO: This looks to be something to do with the determinant of the Jacobian matrix $F_i / x_j$.
\item Intuitively, consider that a patch $dx \cdot dy$ is a slice of a big disk which has dimensions $dr$ on the ray, $r d\theta$ on the arc.
\end{itemize}


\section{3.2: Flux Part I}
Main idea: Field lines are innumerable, so counting them through a surface makes no sense.  Instead, we'll use \textbf{flux} to help us measure charge pushed through a surface per unit time.

Example: If charge $q$ of mass $m$ in a field of $\vec{E} = E_0\hat{i}$ moves from origin along x towards R according to $\frac{d^2x}{dt^2} = \frac{q}{m}E_0$, then solving the diff eq. means that $x = \frac{q}{2m}E_0(\Delta t)^2 = R$.  This means we're pushing all charges within $ \frac{q}{2m}E_0(\Delta t)^2$ to the left of the disk through it.

Then, if we're considering a cylinder of base area $A$, mass density $\delta$, charge density $\rho$:
\begin{itemize}
\item Every test charge chunk $\Delta V$ within $\frac{\rho \Delta V}{2 \delta \Delta V}E_0(\Delta t)^2$ passes through.  That's the height.
\item Area is A, so total volume is  $\frac{\rho (\Delta t)^2}{2 \delta}E_0A$
\item Density of charge per volume is $\rho$, so total is $\frac{\rho^2 (\Delta t)^2}{2 \delta}E_0A$
\end{itemize}

Note: Tilting this forward from the z-axis by $\theta$ multiplies the cross-section area of the cylinder (now an ellipse) by $\cos (\theta)$.  Can work out the ellipse volume, or just note that each ``Riemann bar'' orthogonal to x-axis just got squished by  $\cos (\theta)$. 

So we define \textbf{flux} as amount of charge through a closed surface.  \fbox{$\Phi = (\vec{E} \cdot \hat{n})A$  } if $\vec{E}$ is a constant field.  (Units: joules/second/$m^2$, or watts/$m^2$), and \fbox{$\Phi = \iint_S (\vec{E} \cdot \hat{n})dA$} generally.

We can further note $(\vec{E} \cdot \hat{n}) = \|\vec{E}\| \cos(\theta)$ by last problem.

Example: Flux through an empty cube from the origin is necessarily 0 since every face cancels the other.

Another example: A square pyramind with top at $(0,0,1)$, sides at 1 on each axis:
\begin{itemize}
\item All the triangles will cancel in the x, y directions.
\item A triangle $(1, 0 ,0) (0, 1, 0), (0, 0, 1)$ has two displacement vectors $P_1P_3 = P_3 - P_1 = (-1, 0, 1), P_2P_3 = (0, -1 , 1)$.
\item $P_1P_3 \times P_2P_3 = (1, 1, 1) \rightarrow \hat{n} = \frac{(1, 1, 1) }{\sqrt{3}}$
\item $A = \frac{1}{2} \| P_1P_3 \times P_2P_3 \| = \frac{\sqrt{3}}{2}$
\item $\Phi = (\vec{E} \cdot \hat{n}) A = (E_0 \frac{1}{\sqrt{3}}) \frac{\sqrt{3}}{2} = \frac{E_0}{2}$
\item So total flux through these is $4 \cdot \frac{1}{2} E_0 = 2E_0$
\item However, the bottom has area $\sqrt{2}^2 = 2$ and flux $E_0$ , so total is 0!
\end{itemize}

\section{3.3: Flux Part II}

Note: 
\begin{itemize}
\item Charge ($q$) is the volts of the point charge.  Total charge $Q_{tot}$ is total charge inside some surface.
\item Electric field is sum of those point charges acting at a distance, and a is a single vector.
\item Flux is the sum of the electric field flowing through a surface.
\item Total charge $Q_{tot}$ of a surface is basically the sum of all the flux going in/out, except that it's that divided by some constant $\epsilon_0$. 
\end{itemize}
Note: $\vec{E}$ isn't usually constant, and the surface $S$ is usually curved.  So we need calculus to break up surface $S$ into small pieces $\Delta A_i$ and evaluate $\vec{E}_i$ there at that normal $\hat{n}_i$.  So

$\sum_{patches} (\vec{E}_i \cdot \hat{n}_i) \Delta A_i = \iint_S (\vec{E} \cdot \hat{n})) dA = \Phi$

Easy Example: If, say, $ (\vec{E} \cdot \hat{n}) = 1 $ everywhere, we're just looking at $\iint_S dA$, or the total surface area.

Another example.  Given:
\begin{itemize}
\item Real electric field law: $\vec{E} = \frac{q}{4\pi \epsilon_0} \frac{\vec{x}}{\|\vec{x}\|^3}$
\item Real observation: Total electric flux through a surface ($\Phi$) is proportional to total charge inside ($Q_{tot}$).  $\Phi = \iint_S (\vec{E} \cdot \hat{n}) \Delta A \propto Q_{tot}$
\item Then constant must be $\frac{1}{\epsilon_0} $. Why?
\begin{itemize}
\item On unit sphere, $\hat{n}= \frac{\vec{x}}{\| \vec{x}\|}$
\item So $\vec{E} \cdot \hat{n} = \frac{q}{4\pi \epsilon_0} \frac{\vec{x}}{\|\vec{x}\|^3} \cdot \hat{n}$
\item $= \frac{q}{4\pi \epsilon_0}$ since $\| \vec{x}  \|=1$ on unit sphere
\item Then $\Phi = \iint_S \frac{q}{4 \pi \epsilon_0} dA$
\item  $= \frac{q}{4 \pi \epsilon_0} 4 \pi$ by surface area of unit sphere
\item $= \frac{q}{\epsilon_0}$
\end{itemize}
\item Therefore, because all of the field goes through the surface (no matter the shape), \fbox{\textbf{Gauss's law} says $\iint_S (\vec{E} \cdot \vec{n}) dA = \frac{Q_{tot}}{\epsilon_0}$}
\end{itemize}

\textbf{Note}: Because (UNEXPLAINED!) symmetry of a contained \emph{ball} implies that, for distance $\rho$ from origin, $\vec{E} = E(\rho)\hat{\rho}$, the above works the same for a point charge or a uniform (contained) ball.

Example: For a big radius $R$ ball of charge $Q$ containing a small ball of radius $\rho$ with charge $Q_{tot}$, what must the charge $E(\rho)$ at any point be?

\begin{itemize}
\item Small charge $Q_{tot}$ is proportional to volume of the big charge $Q$ by $Q_{tot} = Q \frac{V_{small}}{V_{big}} = Q\frac{\rho^3}{R^3}$
\item $\frac{Q_{tot} }{\epsilon_0}$ = total charge = $\iint_S E(\rho)(\|\hat{\rho}\|)dA = E(\rho) \iint_S 1 dA  = E(\rho) 4 \pi \rho^2$
\item So $\frac{Q_{tot} }{\epsilon_0} = Q\frac{\rho^3}{R^3 {\epsilon_0}}  = E(\rho) 4 \pi \rho^2$
\item So $E(\rho) = \frac{Q}{4 \pi \epsilon_0} \frac{\rho}{R^3}$
\end{itemize}

Example: Infinite wire, x=y=0, charge per length is $\lambda$.  What's the magnitude of the field r units away?
\begin{itemize}
\item Use a cylinder.
\item What's the total charge of the cylinder?  Top and bottom are perpendicular to the field so can be ignored.
\item There's some function $E(r)$ which, time $\hat{r}$, is the field by symmetry.
\item $\Phi = \iint_{cylinder} (E(r) \cdot \hat{r})dA = E(r)  \iint_{cylinder} 1 dA  = E(r) 2 \pi r h$.
\item $\frac{Q_{tot}}{\epsilon_0} =  E(r) 2 \pi r h \Rightarrow E(r) = \frac{\lambda}{2\pi\epsilon_0 r}$
\end{itemize}


Example: Infinite plane, x=y=0, charge per area is $\sigma$.  What's the mangitude of the field at height $h$?
\begin{itemize}
\item  Use a cylinder again
\item What's the total charge of the cylinder?  Side is perpendicular to the field so can be ignored.  Looking at top and bottom, $\phi = 2EA + 2EA.$, where E is charge through the top.
\item $2EA = \frac{\sigma A}{\epsilon_0} \rightarrow E = \frac{\sigma}{2 \epsilon_0}$
\item Note: It appears it's height-invariant!
\end{itemize}

\section{3.4: Surface Integrals}
\begin{itemize}
\item Flux is a specific form of the general $\iint_S F da$.
\item dA is a patch of a parallelogram on the surface.  This is defined by corners $\vec{x}(u_0, v_0), \vec{x}(u_0, v_0) + \delta_u \vec{x}(u_0, v_0)$, and $\vec{x}(u_0, v_0) + \delta_v \vec{x}(u_0, v_0)$
\item Therefore, using the parallelogram area formula,$dA = \Delta_u \Delta_v \| \vec{x}_u \times \vec{x}_v \|$
\item Taking to the limit, this means the area is $\iint_D F(\vec{x}(u,v)) \|\vec{x}_u \times \vec{x}_v \| du dv$
\end{itemize}

Example: Sphere $x^2 + y^2 + z^2 = R^2$ surface area.  Take $\theta$ as angle around $\phi$ as angle from top of z axis.
\begin{itemize}
\item Parametrization $x = R \sin \phi \cos \theta,  y = R \sin \phi \sin \theta,  z = R \cos \phi$
\item $dx/d_{\theta} =  -R \sin \phi \sin \theta, dy/d_{\theta} =  R \sin \phi \cos \theta, dz/d_{\theta} =  0$
\item $dx/d_{\phi} =  R \cos \phi \cos \theta, dy/d_{\phi} =  R \cos \phi \cos \theta, dz/d_{\phi} =  -R \sin \phi$
\item After working it out, $dx/d_{\theta} \times dx/d_{\phi} = R^2 \sin \phi \langle -\sin \phi \cos \theta,  -\sin \phi \sin \theta, -\cos \phi \rangle$ 
\item Doing the math, $ \|dx/d_{\theta} \times dx/d_{\phi} \| = R^2 \sin \phi $
\item So $\int_{\theta = 0}^{2\pi} \int_{\phi = 0}^{\pi} 1 \cdot  R^2 \sin \phi = 2 \pi \int_{\phi = 0}{\pi} R^2 \sin \phi  = 2\pi R^2 [ -\cos \phi]_0^{\pi} = 4 \pi R^2$
\end{itemize}

Example: Parabaloid $z = 1 - x^2 - y^2, x^2 + y^2 \leq 1$
\begin{itemize}
\item Parametrization $x = R \sin \phi \cos \theta,  y = R \sin \phi \sin \theta,  z = R \cos \phi$
\item $dz/dx = \langle 1, 0, -2x \rangle, dz/dy = \langle 0, -1, -2y \rangle$
\item $\| dz/dx \times dz/dy \| = 1 + 4x^2 + 4y^2$
\item Area = $\iint_D 1 \cdot dA = \iint_D \sqrt{ 1 + 4x^2 + 4y^2} dx dy$
\item Change to polar, remembering this square depends on r: $\int_{\theta = 0}^{2\pi} \int_{r = 0}^{1} \sqrt{ 1 + 4r^2 \cos^2\pi  4r^2 \sin^2\pi} r dr d\theta = 2\pi\int_0^1 \sqrt{1 + 4r^2}rdr$
\item After working it out, this ends up being $[\frac{2}{3} \cdot \frac{1}{8}(4r^2 +1)^{\frac{3}{2}}]_0^1 = \frac{\pi}{6} (5\sqrt{5} - 1)$
\end{itemize}


Example: Torus $x(u,v) = [R+r\cos(u)]\sin(v)], y(u,v) = [R+r\cos(u)]\cos(v)], z = r \sin(u), u, v \in [0, 2\pi)$
\begin{itemize}
\item Already parametrized in polar, basically,
\item $d\vec{x}/du = \langle -r \sin (u) \sin (v),  -r \sin (u) \cos (v), r \cos (u)\rangle$
\item $d\vec{x}/dv = \langle Rcos(v) + r\cos(u)\cos(v), -R\sin(v) -r\cos(v)\sin(v), 0\rangle$
\item After lots of math, $\| d\vec{x}/du \times \vec{x}/dv \| = r(R+ r\cos(u))$
\item $\int_{u = 0}^{2\pi} \int_{v = 0}^{2 \pi}  r(R+ r\cos(u)du = 2 \pi r \int_{u = 0}^{2\pi}r(R+ r\cos(u))du$
\item $= 2\pi r [2 \pi R] = 4 \pi^2 R r$
\end{itemize}


Example: Center of mass of unit (hollow?) hemisphere sitting on origin.
\begin{itemize}
\item Center of mass for density $\rho$ is $\frac{\iint_S \vec{x} \rho dA}{\iint_S \rho dA}$
\item Obvious that $x, y$ center at zero.
\item For denominator, $\iint_S dA$ is just surface area, or half of $4 \pi 1^2 = 2 \pi$.
\item For numerator:
\begin{itemize}
\item Do typical $\theta, \phi$ parametrization.
\item $\vec{x}_{\theta} \times \vec{x}_{\phi} = \langle \sin^2(\phi) \cos(\theta), \sin^2(\phi)\sin(\theta), \sin(\phi)\cos(\phi) \rangle$
\item Pull out the $\sin(\phi)$ and the remaining norm is one, so $\| \vec{x}_{\theta} \times \vec{x}_{\phi}\| = \sin(\phi)$ 
\item $\int_{\theta = 0}^{2\pi} \int_{\phi = 0}^{\pi / 2} z \cdot dA = \int_{\theta = 0}^{2\pi} \int_{\phi = 0}^{\pi / 2} \cos(\phi)\sin(\phi) = \frac{1}{2}$
\end{itemize}

\end{itemize}

Example: Moment of inertia
\begin{itemize}
\item Formula: $I_z = M \iint_S(x^2 + y^2)dA$.
\item Object to spin: helicoid $\vec{x}(\theta,v) = \langle \theta \cos (v), \theta \sin (v), v\rangle \theta \in [0,R], v \in [0, 2\pi]$
\item Assumption for the problem: $\int_{\theta=0}^{\theta=R} \theta^2 \sqrt{1 +\theta^2}d\theta = 2$
\item Center of mass for density $\rho$ is $\frac{\iint_S \vec{x} \rho dA}{\iint_S \rho dA}$
\item Use polar coordinates $r, \theta$.
\item After computation, $\| \vec{x}_r  \times \vec{x}_{\theta} \| = \sqrt{1+r^2}$
\item $M \int_{r=0}^{r=R} \int_{\theta = 0}^{\theta=2\pi}  \sqrt{1+r^2} (r^2 \cos^2(\theta) + r^2 \cos^2(\theta)) dA = M \iint  \sqrt{1+r^2} r^2 = 2\pi M \iint  \sqrt{1+r^2} r^2  = 4 \pi$ by hint
\end{itemize}


Example: Flux through unit hemisphere
\begin{itemize}
\item Formula: $\Phi = \iint_S (\vec{E} \cdot \vec{n}) dA = \iint_S FdA$
\item Field: $\vec{E} = \langle yz, xz, xy \rangle$
\item Use polar coordinates
\item \textbf{Base}: $\hat{n} = -\hat{k}$ so $\langle yz, xz, xy \rangle \cdot \langle 0, 0, -1 \rangle = -xy$ It's clear by symmetry that $\iint_{u^2 + v^2 \leq 1} - xy dx dy = 0$
\item \textbf{Top}: Set $u = \theta \in [0, 2 \pi), v = \phi \in [0, \frac{\pi}{2}]$.  
\item As usual, $dA = \| \vec{x}_u \times \vec{x}_v \| = \sin(v)$.
\item Norm just points out from the center: $\hat{n} = \langle  \cos(u)\sin(v), \sin(u)\sin(v), \cos(v) \rangle $ 
\item $\vec{E} = \langle \sin(u)\sin(v)\cos(v), \cos(u)\sin(v)\cos(v), \cos(u)\sin^2(v)\sin(u)\rangle$ 
\item So $\vec{E} \cdot \hat{n} = 3\cos(u)\sin(u)\cos(v)\sin^2(v)$
\item Looking at this, this is really $\int_{u = 0}^{u = 2\pi} k(v) \sin^2(u)$  for some $k(v)$, so this will be 0.
\item Therefore, total flux is zero, and by Gauss's law, total field contained inside has to be 0 too.
\end{itemize}

Example: Field $\vec{E} = \ln(x^2 + y^2)\langle x, y, 0\rangle$ through $R$-wide cylinder, height $h$
\begin{itemize}
\item Parameterize: $x = r \cos \theta, y = r \sin \theta, z = z$
\item \textbf{Top/Bottom}: $\hat{n} = \langle 0, 0, 1 \rangle, \vec{E} = f(x,y) \langle x, y, 0\rangle \rightarrow \hat{n} \cdot \vec{E} = 0$
\item \textbf{Side}: $\hat{n} = \frac{1}{R} \langle R \cos (\theta), R \sin (\theta), 0 \rangle$
\item $\Phi = \iint_{cylinder} \frac{1}{R} \langle R \cos (\theta), R \sin (\theta), 0 \rangle \cdot  \langle R \cos (\theta), R \sin (\theta), 0 \rangle \ln(R^2 \cos^2(\theta) + R^2 \sin^2(\theta))$
\item $= R \iint_{cylinder} \ln(R^2) + \ln(\cos^2(\theta)+\sin^2(\theta)) = R \cdot 2 \ln (R) \cdot h \cdot 2 \pi R = 4 \pi R^2 \ln(R)h$
\end{itemize}

Example: Field $\vec{E} = e^{-x^2-y^2-z^2}\vec{x}$ with sphere S at radius R, setting $\epsilon_0 = 1$
\begin{itemize}
\item Parameterize: $x = R \cos (\theta)\sin(\phi), y = R\sin (\theta)\sin(\phi), z =R \cos(phi)$
\item $\hat{n} = \langle \cos (\theta)\sin(\phi), y =  \sin (\theta)\sin(\phi), z = \cos(phi) \rangle$
\item $\vec{x} = R \hat{n}$, so $\vec{E} \cdot {n} = R e^{-R^2}$
\item $R \iint_{sphere} e^{-R^2} = 4 \pi R^3 e^{-R^2}$
\end{itemize}

\textbf{Note}: In the future we write \fbox{$\hat{n}dA = \vec{dA}$}

\subsection{3.5: Divergence part I}

\textbf{Main idea}: Last chapter was all about having field $\vec{E}$ and wanting to figure out $Q_{tot}$ (or $\frac{\phi}{\epsilon_0}$). Usually, we have the charge distribution Q and want to figure out $\vec{E}$.  Most of the field derivation from 3.3 was through tricks for highly symmetric spaces (infinite line, infinite plane, uniform ball, etc.)

\textbf{Point}: The flux through a sphere in a uniform field is zero.  Why?  Move the center point to the origin, rotate so field is $\hat{k}$ (both don't change the flux), and consider that what goes out at $\langle x, y, z \rangle$ comes in at $\langle x, y, -z \rangle$.  This same argument applies for $\iint_{S = sphere} \hat{n}_i \hat{n}_j dA$, where $i, j$ are components in $\{x, y, z\}$.


 However, if $i = j$, then $\iint_S \hat{n}_i \hat{n}_j dA = \iint_S \hat{n}^2_i = \frac{4}{3}\pi R^2$, since $\iint_S (\hat{n}_x^2 + \hat{n}_y^2 + \hat{n}_z^2) dA = \iint_S 1 dA = 4 \pi R^2$, so each of the components must be a third of that.

\subsubsection{Defining Divergence}
Remember that, in Gauss's law $\frac{Q}{\epsilon_0} = \iint_S \vec{E} \cdot \vec{dA}$, we're using information about $\vec{E}$ spread out over surface $S$. We can also shrink this to a smaller surface.

Shrinking to a point $\vec{P}$, $\lim_{R \rightarrow 0} \frac{1}{4 \pi R^3}  \iint_S \vec{E} \cdot \vec{dA} = \frac{Q_{tot}}{\epsilon_0 4 \pi R^3} = \frac{\rho(\vec{P})}{\epsilon_0}$. \emph(This works by dividing both sides by volume of a sphere)

\textbf{Deriving Divergence}: Computing $\lim_{R \rightarrow 0} \frac{1}{4 \pi R^3}  \iint_S \vec{E} \cdot \vec{dA}$
\begin{itemize}
\item $\iint_S \hat{n}_i \hat{n}_j dA  = 0$ if $i \ne j$
\item $\iint_S \hat{n}_i \hat{n}_j dA  = \frac{4}{3} \pi R^3 $  if $i = j$
 \item Use linear approximation with Jacobian $D = \frac{\delta E_i}{\delta x_j}$, $\vec{E}(\vec{x}) = \vec{E}(\vec{P}) + D\vec{E}(\vec{P})(\vec{x} - \vec{P})$
 \item $\iint_S \vec{E}(\vec{P}) = 0$ for any constant.  (think of the flux of a sphere in a constant field as above)
\item This leaves $D\vec{E}(\vec{P})(\vec{x} - \vec{P}) \cdot \hat{n} = \sum_{i,j} \hat{n}_i [\vec{x}-\vec{P}]_j D\vec{E}(\vec{P})_{ij}$
\item Since it's a sphere, the normal $\hat{n} = \frac{\vec{x} - \vec{P}}{R}$
\item Therefore $D\vec{E}(\vec{P})(\vec{x} - \vec{P}) \cdot \hat{n} = R \sum \hat{n}_i \hat{n}_j  D\vec{E}(\vec{P})_{ij}$ (swap $R \hat{n}_j$ for $[\vec{x} - \vec{P}]_j$)
\item These terms are all 0 except where $i = j$, so $D\vec{E}(\vec{P})(\vec{x} - \vec{P}) \cdot \hat{n} = \frac{4}{3} \pi R^2 \times R \times
   [\frac{\delta E_x}{\delta x} +  \frac{\delta E_y}{\delta y} +  \frac{\delta E_z}{\delta z}]$
\item This equals $\lim_{R \rightarrow 0} \frac{1}{4 \pi R^3}  \iint_S \vec{E} \cdot \vec{dA}$ so eliminating the sphere volume gives us
\\
   \fbox{$\frac{\rho(\vec{P})}{\epsilon_0} = [\frac{\delta E_x}{\delta x} +  \frac{\delta E_y}{\delta y} +  \frac{\delta E_z}{\delta z}] = \nabla \cdot \vec{E}$}
\end{itemize}

We can think of the divergence  $\nabla$, also like an operator: 
\\
\fbox{$\nabla \cdot \vec{F} = \nabla \cdot (F_x\hat{i} + F_x\hat{j} + F_x\hat{k}) = 
(\frac{\delta}{\delta x} \hat{i} + \frac{\delta}{\delta y} \hat{j} + \frac{\delta}{\delta z} \hat{k}) \cdot (F_x\hat{i} + F_x\hat{j} + F_x\hat{k})$}

\textbf{Shifting to Cylindrical Coordinates}: If instead we want to describe $\vec{F} = \vec{F}_r \hat{r} + \vec{F}_{\theta} \hat{\theta} + \vec{F}_z \hat{z}$, we have 
\fbox{$\nabla \cdot \vec{F} = \frac{1}{r} \frac{\delta r F_R}{\delta r} + \frac{1}{r}\frac{\delta F_{\theta}}{\delta \theta} + F_z \frac{\delta F_z}{\delta z}$}.  How to derive?
\begin{itemize}
\item Note identities $\hat{r} = \cos(\theta) \hat{i} + \sin(\theta) \hat{j}, \hat{\theta} = -\sin(\theta) \hat{i} + \cos(\theta) \hat{j}$.  If $\theta = 0$, these point right and up, corresponding to $\hat{i}, \hat{j}$.  If $\theta$ rotates, these do too.
\item $F_x \hat{i} + F_y \hat{j} + F_z \hat{k} = \vec{F}  = (F_r( \cos(\theta) \hat{i} + \sin(\theta) \hat{j}) + F_{\theta}(-\sin(\theta) \hat{i} + \cos(\theta) \hat{j}) + F_z \hat{k}$
\item Rearrange so that $\vec{F} =  F_x\hat{i} + F_x\hat{j} + F_x\hat{k} = (F_r\cos(\theta)   + F_{\theta}(-\sin(\theta)) \hat{i} +( F_r \sin(\theta) + F_{\theta}\cos(\theta))\hat{j} + F_z\hat{k}$.
\item Compute $\frac{\delta}{\delta x} = \frac{\delta}{\delta r} \frac{\delta r}{\delta x}  + \frac{\delta}{\delta \theta} \frac{\delta \theta}{\delta x} =
\frac{\delta}{\delta x} = \cos(\theta)  \frac{\delta}{\delta r}  - \frac{\sin(\theta)}{r} \frac{\delta}{\delta \theta} $.  
\begin{itemize}
\item The second term: $\frac{d\theta}{dx} = \tan^{-1} (y/x)) = \frac{y}{1+y^2/x^2} * \frac{-1}{ x^2} = -\frac{r\sin(\theta)}{r^2(\sin^2 + \cos^2)} = -\frac{\sin (\theta)}{ r}$
\end{itemize}
\item Do something similar for similar for $\frac{d}{dy}$ in the second term.
\item Combine and shake it out.
\end{itemize}



\textbf{Shifting to Spherical Coordinates}: Using a similar process, we get 
\\
\fbox{
$\nabla \cdot \vec{F} = \frac{1}{\rho^2} \frac{\delta (\rho^2 F_{\rho})}{\delta\rho} + \frac{1}{\rho \sin(\phi)} \frac{\delta}{\delta \phi} (\sin(\phi)F_{\phi}) + \frac{1}{\rho \sin(\phi)} \frac{\delta F_{\theta}}{\delta \theta}$}




\subsection{3.6:  Divergence Part 2}

Example: Compute divergence of electric field $E = \frac{Q}{4 \pi \epsilon_0}\frac{\vec{x}}{\| \vec{x} \|^3}$ outside radius R.
\begin{itemize}
\item $\frac{\delta E_x}{\delta x} (\frac{Q}{4 \pi \epsilon_0}\frac{x}{(x^2 + y^2 + z^2)^\frac{3}{2}}) = v\frac{-2x^2 + y^2 + z^2}{(x^2 + y^2 + z^2)^\frac{5}{2}}$.
\item Symmetrical for $\frac{\delta E_y}{\delta y}, \frac{\delta E_z}{\delta z}$
\item Sums to 0.
\end{itemize}

Example: Compute divergence of electric field $E = \frac{Q}{4 \pi \epsilon_0}\frac{\vec{x}}{R^3}$ inside radius R.
\begin{itemize}
\item $\frac{\delta E_x}{\delta x} (\frac{Q}{4 \pi \epsilon_0}\frac{x}{R^3}) =  \frac{Q}{4 \pi \epsilon_0}\frac{1}{R^3}$
\item Symmetrical for $\frac{\delta E_y}{\delta y}, \frac{\delta E_z}{\delta z}$
\item Sums to $\frac{Q}{4 \pi \epsilon_0}\frac{3}{R^3}$
\end{itemize}

So, the divergence of an electric field is proportional to $\frac{Q}{R^3}$ inside the sphere, and 0 outside the sphere.  

So, divergence at a point intuitively measures \textbf{how much the field spreads out} or sinks into the point.  For electric charge, 
\fbox{$\nabla \cdot \vec{E} = \frac{\rho}{\epsilon_0}$} means that at that point, the spready-ness is proportional to the charge.

Example: if $\epsilon_0 = 1$ and the field is $\vec{E} = x \hat{i} + 2y \hat{j} + z \hat {k}$, how much charge is in the $[0,1] \times [0,1] \times [0,1]$ box?
\begin{itemize}
\item	Answer: $\rho = \nabla \cdot \vec{E} = 1 + 2 + 1 = 4$.  So 4 units.
\end{itemize}


Another Example: if $\vec{E} = \sin(yz) \hat{i} + \sin(xz) \hat{j} + \sin(xy) \hat {k}$ in some complicated surface, then what?
\begin{itemize}
\item	Noticing that $\nabla \cdot \vec{E} = 0$ shows you this is 0 no matter the shape of the region.  This means \emph{the vectors pointing into the region (in fact, any part of the space) are balanced out by those pointing out from the region.}
\end{itemize}


\subsection{3.7:  The Divergence Theorem}

\textbf{The Divergence Theorem} falls out of equating finding charge $Q$ with a double integral over a bounded surface with the triple integral of the contained volume:
\begin{itemize}
\item $\frac{Q}{\epsilon_0} = \iint_S \vec{E} \cdot \vec{dA}$ (Gauss's law)
\item $\Rightarrow \nabla \cdot \vec{E} = \frac{\rho}{\epsilon_0}$ within R (Proved Divergence equivalent from last section)
\item $Q = \iiint_R \rho dx dy dz$ (Just integrating charge over volume)
\item $\Rightarrow Q = \iiint_R \rho dx dy dz = \epsilon_0  \iiint_R \nabla \cdot \vec{E} dx dy dz = \epsilon_0 \iint_S \vec{E} \cdot \vec{dA}$ 
\item \fbox{$\Rightarrow   \iint_S \vec{E} \cdot \vec{dA} = \iiint_R \nabla \cdot \vec{E} dx dy dz$} (Divergence Theorem).  Integrating the field over the bounding surface is the same as integrating the divergence over the volume.
\end{itemize}


\textbf{Smooshy thought}: This looks like another version of Fundamental Theorem of Calculus.  The integral of the function evaluated at the boundaries is the same as the function summed inside the boundary.


\textbf{Proving Divergence Generally}: We're gluing micro-cubes together and not changing the total flux.  This means any surface is the flux going in and out of its "skin".

\begin{itemize}
\item Note that since the flux outward through a cube face is the negative of it inward, gluing two cubes together on this face means we're summing the total fluxes.
\item Do this for tiny cubes approximating the surface we care about.
\item In a cube centered on point $P, F \approx \vec{F}(P) + D\vec{F}(P)(\vec{x} - \vec{P})$.
\item $\iint_S \vec{F}(P)\vec{dA} = 0$ since it's constant, since every face $i$ has normal $\hat{n}_i$, and a partner of equal size with normal $-\hat{n}_i$.
\item However,  for a cube of side $\epsilon$ the flux through, say, Face I ($x = \epsilon + P$) is  $\iint_S  D\vec{F}(P)(\vec{x} - \vec{P})$ ends up being $\frac{\delta F_x}{\delta x} 4 \epsilon^3$, since:
\begin{itemize}
\item Consider side $x = p_x + \epsilon$
\item $D\vec{F}(P)(\vec{x} - \vec{P}) \cdot \hat{n} =  [D\vec{F}(P)]_{xx} (x - p_x) + [D\vec{F}(P)]_{xy} (x - p_y) + [D\vec{F}(P)]_{xz} (z - p_z)$.
\item So, the functons that consider the inputs of y, z don't matter.
\item So $\iint_{Face I} (y - p_y)dA = 0$  around $p_y$ by symmetry.  Same for z on that face.
\item But for $x$, $\iint_{Face I} (x - p_x)dA = \int_{p_y - \epsilon}^{p_y + \epsilon} \int_{p_z - \epsilon}^{p_z + \epsilon}  \epsilon dy dz = 4 \epsilon^3$ 
\item $D_{ij}\vec{F}(P)$ is constant for all $i, j \in \{x, y, z\}$, so this face is then $\frac{\delta F_x}{\delta x} 4 \epsilon^3$.
\item Summing the opposite face (with the same flux), yields $\frac{\delta F_x}{\delta x} 8 \epsilon^3$ = $\frac{\delta F_x}{\delta x} V$.
\item Summing across the other faces yields $\frac{\delta F_x}{\delta x} V + \frac{\delta F_y}{\delta y} V + \frac{\delta F_z}{\delta z} V$.
\end{itemize}

Finally, this shows the \textbf{flux on one of these microcubes is} \fbox{ $\nabla \cdot \vec{F}(P) V$.}

In total, the \textbf{divergence theorem}: \fbox{ $\iint_{\delta C} \vec{F} \cdot \vec{dA} \approx \nabla \cdot \vec{F}(P) V \approx \iiint_C \nabla \cdot \vec{F} dx dy dz$}

\end{itemize}


\textbf{Example of using divergence to calculate flux}: Unit hemisphere with $\vec{E} = \langle yz, xz, xy \rangle$:
\\
\emph{Answer}: $\nabla \cdot \vec{E} = \frac{\delta}{\delta x} yz +  \frac{\delta}{\delta y} xz +  \frac{\delta}{\delta z} xy  = 0$

\textbf{Example of using divergence to calculate flux}: Cylinder of radius $R$, height $h$, sitting on $z = 0$ with $\vec{E} = \ln(x^2 + y^2) \langle x, y, 0 \rangle$:
\\
Answer:
\begin{itemize}
\item $\frac{\delta}{\delta x} E_x =  \ln(x^2 + y^2) + \frac{2x^2}{x^2 + y^2}$.  Similar for $E_y$.
\item Transform to polar: $E_x + E_y = \ln(r^2) + \frac{2 r^2 \cos(\theta)^2 + r^2 \sin(\theta)^2}{r^2} = 2\ln(r) + 2$
\item Set up the integral, remembering the Jacobian: $\Phi = 2 \pi \int_{z = 0}^h \int_{r = 0}^{r=R} [ 2\ln(r) + 2 ] r dr d\theta$
\item Working it out, with identity $\int x \ln(x) = -\frac{x^2}{4} + \frac{x^2}{2}ln(x)$, you get $\Phi = 4 \phi R^2 ]\ln(R) h$
\end{itemize}

\textbf{Example of using divergence to calculate flux}: Unit sphere at origin with $\vec{E} = (x^3 + y^3)\hat{i} +  (z^3 + y^3)\hat{j} +  (x^3 + z^3)\hat{k}$ 
\\
Answer:
\begin{itemize}
\item $ E_x + E_y + E_z =  3x^2 + 3y^2 + 3z^2  = 3 \iiint \rho^2 dx dy dz$
\item Each $d \rho$ is a sphere of volume $4 \pi f(\rho)^2 = 4 \pi \rho^4$
\item So the integral is $12 \pi \int_{\rho = 0}^{\rho = 1} \rho^4 = \frac{12\pi}{5}$ 
\end{itemize}


\textbf{Example of using divergence to calculate flux}: $\vec{F} =  (\cos(z)+x^2)\hat{i}, + (xe^{-z})\hat{j} + (\sin(y) +x^2z)\hat{k}$  on parabaloid $z = x^2+y^2, z \leq 4$ with top $x^2+y^2 \leq 4, z=4$
\begin{itemize}
\item $\iiint_R \nabla \cdot \vec{E} = \int_{z = r^2}^4 \int_{x^2 + y^2 = 0}^2 (y^2 + x^2) dx dy dz$
\item $\iiint_R \nabla \cdot \vec{E} = \int_{\theta = 0}^{2 \pi} \int_{r=0}^2 \int_{z=r^2}^4 (r^2\cos^2(\theta) + r^2\sin^2(\theta)) r d\theta dr dz = r^3 d\theta dr dz$
\item $= 2 \pi  \int_{r=0}^2 4r^3 - r^5 = 2\pi[r^4 - \frac{r^6}{6} ]_0^2 = \frac{32}{3}\pi$.
\end{itemize}

\textbf{What's crazy}: Evaluating divergence of a point charge $\vec{E} = \frac{Q}{4 \pi \epsilon_0}\frac{\vec{x}}{\|\vec{x}\|^3}$
\begin{itemize}
\item $\frac{\delta}{\delta x} E_x = \frac{Q}{4 \pi \epsilon_0} \frac{\delta}{\delta x} x (x^2+y^2+z^2) = \frac{-2x^2 + y^2 + z^2}{(x^2 + y^2 + z^2)^\frac{5}{2}}$
\item $E_y, E_z$ follow symmetrically.
\item The sum is infinite at the origin and zero everywhere else
\item Therefore, they had to invent a $\delta$ function that is infinite at origin, 0 elsewhere, and $\iiint_{\mathbb{R}^3} \delta({\vec{x}}) = 1$
\end{itemize}

\subsection{3.8: Divergence and Fluids}

Looking back to section 3.1, this hydrostatic force function should follow similar patterns to flux: \fbox{$\vec{F}_{tot} = \iint_S p \hat{n} dA$}.

Extended Example: A round ball of radius $R$, center at depth $h$ , with force $\vec{F}_{tot} = p_0 \iint_S [1 - \frac{z}{h}] \hat{n} dA$.
\begin{itemize}
\item $\hat{n} = \frac{\langle x, y, z \rangle}{R}$
\item For the integral, note that x, y are completely symmetric around z axis, so they contribute 0.
\item For the integral, we're then looking at  $\frac{p_0}{R} \iint_S [1 - \frac{z}{h}]zdA$
\item Use spherical coordinates: $\frac{p_0}{R} \iint_S [1 - \frac{R\cos(\phi)}{h}] R\cos(\phi)dA$
\item Working through $dA = \sqrt{1 + f_x^2 + f_y^2}dx dy$ with $f = R - \sqrt{R^2 + x^2 + y^2}$, we get $dA = \frac{dxdy}{\sqrt(1 - \frac{x^2 + y^2}{R^2})}$
\item This $dA$ term, in spherical coordinates, becomes $R^2 \sin(\phi)d\phi d\theta$
\item Combining and substituting $u = \cos(\phi)$, this integral is $-\frac{4\pi R^3}{3}\frac{p_0}{h}\hat{k}$
\end{itemize}

The neat idea: $F_{tot} = \frac{4\pi R^3}{3} \times -\frac{p_0}{h}\hat{k}$ is really ``ball's volume'' times a constant.
\begin{itemize}
\item $\frac{4\pi R^3}{3} \times  -\frac{p_0}{h}\hat{k}$
\item $= \iiint_B 1 dx dy dz  \times  -\frac{p_0}{h}\hat{k}$

\item $= \iiint_B( -\frac{p_0}{h})\hat{k} dx dy dz$, with $p = p_0[1-\frac{z}{h}]$
\item $=  \iiint_B(\frac{\Delta p}{\delta z}) \hat{k} dx dy dz$
\item $=  \iiint_B \nabla \cdot p dx dy dz$
\item So the upshot is the divergence theorem again:\fbox{ $\iint_{S} p \hat{n}dA = \iiint_B \nabla \cdot p dx dy dz = \iiint_B \nabla \cdot p d\vec{x}$}
\end{itemize}

Final example three ways: ``oxygen flow'' (really, flux) through ball of radius R at origin, under field $J = j_0\hat{i}$.
\begin{itemize}
\item Intuitive: what comes in at (-x, y, z)  goes out at (x, y, z), so total is zero.
\item Flux integra under spherical: $\iint \vec{F} \hat{n} dA = \iint_S j_0\hat{i} \cdot \frac{\langle x,y,z \rangle }{R} dA = \int_{\theta = 0}^{2\pi}\int_{\phi = 0}^{\pi} j_0 \cos(\theta)\sin(\phi) R^2 \sin(\theta) d\theta d\phi = 0 = \frac{\pi j_0 R^2}{2} \int_{\theta = 0}^{2\pi} \cos(\theta) d\theta = 0$
\item Divergence: $\nabla \cdot \vec{J} = \frac{\delta}{\delta x} j_0 + 0 + 0 = 0, so \iiint_B 0 = 0$.
\end{itemize}

\subsection{3.9: Flows and Divergence}

Main idea: Divergence ($\nabla \cdot \vec{V}$) measures how much the flow changes volumes at that point.

Example: What is the function described by field of velocity vectors $\vec{V}(\vec{x}) = \langle -y, x \rangle$?
\begin{itemize}
\item $x'(t) = -y, y'(t) = x$
\item $\Rightarrow x''(t) = -y' = -x, y''(t) = x' = -y$
\item $\Rightarrow x = A\cos(t) + B\sin(t), y = C\cos(t) + D\sin(t)$, work it out to $x = \cos(t), y = \sin(t)$
\end{itemize}

Idea: dump a $dA=s_1$ by $s_2 =  \Delta x\hat{i} \times \Delta y\hat{j} $rectangle into the flow and see how it deforms over time.  Over a long time, it'll distort a lot, but consider for $\Delta t$:
\begin{itemize}
\item dA has sides of length $\Delta x, \Delta y$ but area of dA: $\| s_1 \times s_2\|$ (cross product norm is parallelogram area)
\item What is side $s_1$ after $\Delta t$? The starting point plus (how the endpoint moves minus how the start point moves): $\vec{s_1} + \Delta t[\vec{V}(x_0 + \Delta x, y_0) - \vec{V}(x_0, y_0)]$
\item Expanding the iterated $s_1$, which we call $s_1'$ out: $\vec{s_1}' = \Delta x[(1 + \Delta t\frac{\delta V_x}{\delta x})\hat{i} + \Delta t \frac{\delta V_y}{\delta x} \hat{j}]$.  Do the same for $s_2'$ and work out in 3D: $s_1' \times s_2' \approx \Delta x \Delta y [\hat{k} + \Delta t[(V_x)_x + (V_y)_y]\hat{k}$
\item We end up with $s_1' \times s_2' \approx \Delta x \Delta y [1 + \Delta t \nabla \cdot \vec{V}]\hat{k}$, so vs. original area $\Delta x \Delta y$, the ratio is $1 + \Delta t \nabla \cdot \vec{V}$
\item This means \textbf{divergence $\nabla \cdot \vec{V}$ is proportional to the change in area due to the flow}!
\end{itemize}

An \textbf{incompressible} field preserves volume under flow (so $\nabla \cdot \vec{V} = 0)$, like $\langle y, z, x\rangle, \langle 0, 2\sqrt{x}, 0 \rangle,  \langle x, y, -2z \rangle$.

A cool interactive on the page shows how a sphere migrating its points via $\langle x, y, z \rangle$ grows and changes volume, while one under $0.3 \langle  y, z , x \rangle$  distorts but doesn't.

\section{Chapter 4: Work, Line Integrals, Stokes's Theorem}
\subsection{4.1: Work Part I}

\textbf{Energy U} for charge $q$ in field $\vec{E}$ is \fbox{$-q\vec{E} = \nabla U$}.
\\
\emph{Example}: Right-pointing constant field $\vec{E} = E_0\hat{i}$ means $(\frac{\delta}{\delta x}\hat{i} + \frac{\delta}{\delta y}\hat{j} + \frac{\delta}{\delta z}\hat{k})U = -qE_0\hat{i} \Rightarrow U = -qE_0x$

\textbf{Work} is change in energy, e.g. $W_{field} = -[U(\vec{x_f} )- U(\vec{x_0})]$.  If the charge flows with the field, then the field is doing positive work.  If the charge flows against the field, the field is doing negative work.  So in the above case, $W_{field} = -[U(x_f,0,0) )- U(x_0, 0, 0)]  \Rightarrow -[-qE_0x_f - -qE_0x_0 = qE_0(x_f - x_0)$.

Continuing the example, moving from $(x_0,0, z_0) $ to $(x_f,0, z_f)$ in $\vec{E} = E_0\hat{i}$: only the x-coordinate affects the energy, so $U(x) = -qE_0x$ and $W_{field} = qE_0(x_f - x_0)$.  If $s$ is the distance between the two, then $\cos(\theta) = \frac{|x_f - x_0|}{s} \Rightarrow W_{field} = qE_0 s \cos(\theta)$.

Expanding the example, consider $\vec{E}(\vec{x}) = \vec{E_0}$, a constant that may not be aligned just with x-axis.  Then $-q\vec{E} = \nabla U \Rightarrow \nabla (-q\vec{E_0} \vec{x} + C) = -q\vec{E} \Rightarrow U = -q \vec{E_0}\vec{x} +C.$ Can also solve the diff eq, more generally.  This also means that, still, $W_{field} = q\vec{E_0}(\vec{x_f} - \vec{x_0})$.

Big idea: Though most fields aren't constant, they are near-constant between a small displacement $\Delta x$.  
\begin{align}
W_{field} = q\vec{E} \cdot \Delta \vec{x} \\
= \sum q\vec{E}(\vec{x}(t_i)) \cdot [\vec{x}(t_{i+1}) - \vec{x}(t_i)] \\
= \sum q\vec{E}(\vec{x}(t_i)) \frac{\Delta x(t_{i+1})} {\Delta t}\Delta t 
= \int_a^b q \vec{E}(\vec{x}(t)) \cdot \frac{d\vec{x}}{dt}dt
\end{align}

So we can take a \textbf{line integral} to measure work over a path in a field.

\subsection{4.2: Work Part 2}

In general, the \textbf{work} for moving a charge $q$ along path $\vec{x}(t)$ through field $\vec{E}(\vec{x})$ is \fbox{$W = \int_a^b q\vec{E}(\vec{x}(t)) \cdot \frac{d\vec{x}}{dt} dt$}.  More generally, $q\vec{E}$ is just a force, so we're looking at \fbox{$W_{field} = \int_a^b \vec{F}(\vec{x}(t)) \cdot \frac{d\vec{x}}{dt} dt$}

Example: $\vec{x}(t) = \langle t+1, t+1, t \rangle, t \in [0, 8], \vec{E}(x,y,z) = \frac{\lambda}{2 \pi \epsilon_0} \frac{x \hat{i} + y \hat{j}}{x^2+y^2}$
\begin{align}
W = \frac{q \lambda}{2 \pi \epsilon_0} \int_{t=0}^8 \frac{\langle t+1, t+1, t\rangle \cdot \langle 1, 1, 0 \rangle}{2(t+1)^2} \\
=\frac{q \lambda}{2 \pi \epsilon_0} \int_{t=0}^8 \frac{1}{t+1} \\
= \frac{q \lambda}{2 \pi \epsilon_0} \int_{u=1}^9 \frac{1}{u}  \\
= \frac{q \lambda}{2 \pi \epsilon_0} \ln(9)
\end{align}

However, what happens if we keep the endpoints but change the path?

Extended example: $\vec{x}(t) = \langle \sqrt{2}t \cos(\frac{\pi t}{4}), \sqrt{2}t \sin(\frac{\pi t}{4}), t-1\rangle$, still with $\vec{E}(x,y,z)= \frac{\lambda}{2 \pi \epsilon_0} \frac{x \hat{i} + y \hat{j}}{x^2+y^2}$.  
\begin{align}
\vec{E}(\vec{x}(t))  \cdot \frac{d\vec{x}}{dt} = ... = 2t \\
x^2 + y^2 = 2t^2 \\
W =  \frac{q \lambda}{2 \pi \epsilon_0} \int_{1=1}^9 \frac{1}{t}  \\
= \frac{q \lambda}{2 \pi \epsilon_0} \ln(9)
\end{align}

So it looks like there \emph{might be path independence here}.  Consider: rubbing your hands together - is there more work done oscillating and ending at the initial position than doing nothing? (Clearly yes).

Example: Work due to friction. $\vec{F} = -\gamma \frac{d\vec{x}}{dt}, \gamma > 0$.  Move from $(0,0,0) \rightarrow (1,0,0)$ via $\vec{x}(t) = \langle t, 0, 0 \rangle$.  If $\gamma = 1$, what is $\int_a^b \vec{F}(\vec{x}(t)) \cdot \frac{d \vec{x}}{dt} dt$?

Answer: $-\int_{t=0}^1 \frac{d \vec{x}}{dt}^2dt = - \int_{t=0}^1 \langle 1, 0, 0 \rangle ^2 = -1$  We lose 1 unit of energy.

To illustrate that there's not path independence always, consider an oscillating object following $\vec{x}(t) = \langle t, \sin(n\pi t), \sin(n \pi t) \rangle, t \in [0,1]$.  Take $\gamma = 1$ so $\vec{F} = \vec{x}'(t).$

Answer: $- \int_{t=0}^1 \vec{x}'(t) \cdot \vec{x}'(t) dt = - \int_{t=0}^1 (1+2n^2\pi^2\cos^2(n \pi t)) dt =  -\int_{t=0}^1(1+ 2n^2\pi^2 \frac{1}{2}(1 + \cos(2 n \pi t))  dt = -\int_{t=0}^1 (1 + n^2\pi^2+2n^2\pi^2\cos(2n\pi t))dt = -[ t + tn^2\pi^2 =+n\pi \sin(2 n \pi t)]_0^1 = -[1 + n^2\pi^2]$

Example: Spring force $\vec{F} = -\frac{\|x - l_0\|}{\|x\|} \vec{x}$ along path $\vec{x}(t) = \langle 0, 1-t, 2t\rangle, t \in [0,1]$

Answer: $\int_{t=0}^1 \vec{F}(\vec{x}(t)) \frac{d \vec{x}}{dt}dt = - \int_{t=0}^1 \frac{\sqrt{5t^2 - 2t + 1} - 1}{\sqrt{5t^2 - 2t + 1}} \langle 0, 1-t, 2t \rangle \cdot \langle 0, -1, 2 \rangle= \int_0^1 [1 - (5t^2 - 2t + 1)^\frac{-1}{2}](-1 + 5t)dt = -[-t + \frac{5}{2}t^2 - (5t^2 - 2t +1)^\frac{1}{2}]_0^1 = -\frac{1}{2}$

Example: \emph{SAME} Spring force $\vec{F} = -\frac{\|x - l_0\|}{\|x\|} \vec{x}$ and \emph{SAME} endpoints but along path $\vec{x}(t) = \langle 0, \cos(t), 2\sin(t) \rangle,  t \in [0,1]$

Answer: $\int_{t=0}^{\pi / 2} \vec{F}(\vec{x}(t)) \frac{d \vec{x}}{dt}dt  = - \int_{t=0}^{\pi / 2} (1 - \frac{1}{\|x\|})\langle 0, \cos(t), 2\sin(t) \cdot \langle 0, -\sin(t), 2\cos(t) \rangle= -3  \int_0^{\pi / 2} \sin(t)\cos(t)(1 - \frac{1}{\sqrt{1+3t^2}}dt = 3 \int_{t=0}^{\pi / 2} - \sin(t)\cos(t) + 3 \int_{t=0}^{\pi / 2} \sin(t)\cos(t) \frac{1}{\sqrt{1+3\sin^2(t)}} = 3[\frac{1}{2}\cos^2(t)]_0^{\pi / 2} + 3[\frac{2}{6}(1+3\sin^2(t))^{1/2}]_0^{\pi / 2} = -1/2$

So sometimes path does not matter (electricity and spring) but sometimes it seems to matter (friction). Something about the ``swirl'' of the field (curl)?  Addressed upcoming.


\subsection{4.3: Line Integrals}

A line integral is 
\begin{itemize} 
\item Most generally: $\int_{s=0}^{s=l} f(\vec{x}(s))ds$ or $\int_C f ds$.
\item Built out of function heights $h_i$ over curve snippet lengths $\Delta s_i$ : $\sum_i h_i \Delta s_i$.
\item More practically written as $\sum f(x_i, y_i)\sqrt{[\Delta x]^2 + [\Delta y]^2}$
\item Riemanned up: $\int_{s=0}^{s=l} f(\vec{x}(s))ds$
\item Practical integral based off of some $t: \int_C f(\vec{x}(t))\|\vec{x}'(t)\|dt$
\end{itemize}

Example: Area of curtain, height = $f(x,y) = y^2$, base is origin circle of radius 2 on xy-plane.
\begin{itemize}
\item $A = \int_{t=0}^{t=2\pi} f(x,y) \|  \frac{d<2 \cos (t), 2 \sin (t), 0>}{dt} \|dt$
\item $A = \int_{t=0}^{t=2\pi} 4\sin^2(t)  * 2 = 4 \int_{t=0}^{t=2\pi} (1 - \cos(2t)) = 4[t - \frac{\sin(2t)}{2}]_0^{2\pi} = 8 \pi$
\end{itemize}

Example: Geometric area of $f(x,y)=4y^3$ over the curve $x = \frac{y^3}{3}, (\frac{-1}{3}, 1) \rightarrow  (\frac{1}{3}, 1)$

\begin{itemize}
\item $\vec{x}(t) = \langle t^3 / 3, t \rangle \Rightarrow \vec{x}'(t)  = \langle t^2, 1 \rangle \Rightarrow \| \vec{x}'(t)  \| = \| \sqrt{1 + t^4} \|$
\item Note that this is an odd function: the part where $y \in [-1, 0]$ is a negative of [0,1].  So we'll double the right half.
\item $A = 2 \int_{t=0}^{t=1} 4t^3 \sqrt{1+t^4}dt =[\frac{4}{3}(1+t^4)^\frac{3}{2}]_0^1 = \frac{4}{3}[2\sqrt{2}- 1]$
\end{itemize}

Example: Moment of interatia around z-axis: $\int[x^2+y^2]\rho ds$ for $\vec{x}(t)=\langle 2\sin(t), 2\cos(t), 3t\rangle t \in [0,2\pi]$ if $\rho = \frac{1}{2\pi\sqrt{13}}$:
\begin{itemize}
\item $\vec{x}'(t) = \langle 2\cos(t), 2\sin(t), 3 \Rightarrow \| \vec{x}'(t)  \| = \sqrt{4 + 9}$
\item $x^2 + y^2 = (2\sin(t))^2 + (2\cos(t))^2 = 4$
\item $A = \rho * \int_{t=0}^{t=2\pi} 4 \sqrt{13} =  \frac{1}{2\pi\sqrt{13}} * 2\pi * 4\sqrt{13} = 4$
\end{itemize}

Example: Infinite wire of current going up on the x-axis.  Field equation, for (generated?) field $\vec{B}$, penetrated region bounded by curve $C$, unit tangent (to the curve) $\hat{T}$ , constant $\mu_0$ and current amount $I$: \fbox{$\int_C [\vec{B} \cdot \hat{T}]ds = \mu_0I$}.  What is $\| \vec{B} \|$ at some distance from the z-axis $r$?
\begin{itemize} 
\item C should be a circle of radius $r$.  Then, $ds = \frac{d\vec{x}}{dt}dt = r \| \cos(t), \sin(t) \| = r$
\item $\vec{B} = \|B\| \hat{T}$ by definition I suppose.
\item $\int_{t = 0}^{t=2\pi} [ \|B\| \hat{T} \cdot \hat{T}] r dt$
\item $\|B\| * 2\pi r = \mu_0 I$
\item $\Rightarrow \|\vec{B}\| = \frac{\mu_0 I}{ 2\pi r }$

\end{itemize} 

\subsection{4.4: Path independence}

When do we care only about the endpoints and not the path?  Examples include the spring and the electric charge, and counterexamples include motion under friction.

Note that integrals of the form $\int_C \frac{df}{ds}ds = \int_{s=0}^{s=l} f'(s)ds = f(x(s=l)) - f(x(s=0))$ are path independent (fundamental theorem of calculus).  \textbf{So if you can find a legit antiderivative of our function, it is path independent.}  Or, if $\vec{F}$ is a \textbf{gradient vector field} ($\vec{F} = \nabla f$), then the line integral depends only on the endpoints of C.

Note: If $f(s) = f(\vec{x}(s))$, then $\int_C f'(s) = \int_C f'(\vec{x}(s)) = \int_{s=0}^{s=l} \frac{\delta f}{\delta x} (x(s))x_x'(s) + \frac{\delta f}{\delta y} (x(s))x_y'(s) + \frac{\delta f}{\delta z} (x(s))x_z'(s)$ or \fbox{$\int_{s=0}^{s=l} \nabla f (x(s)) \cdot x'(s)ds = \int_C f'(s)$}.  \emph{This looks less like a magical theorem and more of a description of a componentwise derivative, line-integrated over ds like usual.}

Example: if $\vec{E} = \frac{\lambda}{2\pi\epsilon_0}\frac{x\hat{i} + y \hat{j}}{x^2 + y^2}$, find $f$ such that $\nabla f = \vec{E}$.
\begin{itemize}
\item Note: This is just the anti- and derivative cycle technique.
\item What's on the $\hat{i}$?  That's $\frac{\delta f}{\delta x} = \frac{\lambda}{2\pi\epsilon_0}\frac{x}{x^2 + y^2}$
\item What's on the $\hat{j}$? That's $\frac{\delta f}{\delta y} = \frac{\lambda}{2\pi\epsilon_0}\frac{y}{x^2 + y^2}$
\item Anti-derivative on the x: $f = \frac{\lambda}{4\pi\epsilon_0}\ln(x^2+y^2) + h(y)$
\item Re-differentiate for y: $\frac{\delta f}{\delta y}  = \frac{\lambda}{2\pi\epsilon_0}\frac{y}{x^2 + y^2} + h'(y)$
\item So $h(y)$ is constant, and $f = \frac{\lambda}{4\pi\epsilon_0}\ln(x^2+y^2) + C$
\end{itemize}

Example: Another cycle:  if $\vec{F} = (2x - 3y)\hat{i} + (6y - ax)\hat{j}$, what's an $a$ to allow for $\nabla f = \vec{F}$?
\begin{itemize}
\item Note: This is just the anti- and derivative cycle technique again
\item What's on the $\hat{i}$?  That's $\frac{\delta f}{\delta x} = 2x-3y$
\item What's on the $\hat{j}$? That's $\frac{\delta f}{\delta y} = 6y-ax$
\item Anti-derivative on the x: $f = x^2 - exy + h(y)$
\item Re-differentiate for y: $\frac{\delta f}{\delta y} = -3x + h'(y) = -3x + 6y$.  $a = -3$.
\end{itemize}

Example: Spring:  if $\vec{F} = -(\|\vec{x}\| - l_0)\frac{\vec{x}}{\| \vec{x} \|}$, find $f$ such that $\nabla f = \vec{F}$.
\begin{itemize}
\item Hint given: $\nabla(\frac{1}{2} \|\vec{x}\|^2) = \vec{x} \Rightarrow \nabla (\| \vec{x} \|) = \frac{\vec{x}}{\|\vec{x}\|}$
\item Rewrite as $-\vec{x} + l_0 \frac{\vec{x}}{\|\vec{x}\|}$
\item KEEP the nablas!
\item Use hints on each term: $\vec{F} = -\nabla(\frac{1}{2} \|\vec{x}\|^2) + l_0\nabla(\|x\|)$
\item $\vec{F} = \nabla(-\frac{1}{2} \|\vec{x}\|^2 + l_0\|x\|)$
\item Note: Second term worked out through chain rule, or you can just do the hand-math on $\sqrt{x^2 + y^2...}$
\item \textbf{The magic}: Now we can use $f$ to find the work from earlier problems: $f(0, 0, 2) - f(0, 1, 0)  = -\frac{1}{2}$!
\end{itemize}

\textbf{Curve orientation}: pick two opposite parametrizations of the curve $C_+$ and $C_-$ (say, ``going right'' and ``going left'').  We can use right-hand rule, where, if curve in some xy plane, right thumb points along $\hat{k}$ and $C_+$ curls with fingers.

We know that $\int_{C_+}\vec{F}\cdot d\vec{x} = -\int_{C_-}\vec{F}\cdot d\vec{x}$ and therefore, $\int_{C_+ + C_-}\vec{F}= 0$

We can say \fbox{If $\int_C \vec{F}\cdot d\vec{x}$ depends only on endpoints of $C$ then $\vec{F}$ is a gradient field.}
\begin{itemize}
\item Basic idea: Similar to the fundamental theorem $g(x) = \frac{d}{dx} \int_{t=a}^{t=x}g(t)dt$, we can prove $\vec{F}(\vec{u}) = \nabla [\int_{\vec{a}}^{\vec{u}} \vec{F} \cdot d\vec{x} ]$.
\item Plan: prove directional derivative of $\int \vec{F} \cdot d\vec{x}$ along $\vec{v}$ is $\vec{v} \cdot \vec{F}(\vec{u})$ for any $\vec{v}$, which proves the answer is $\vec{F}(\vec{u})$
\item Neat idea: We're looking for  $\int_{\vec{a}}^{\vec{u}+h\vec{v}} \vec{F} \cdot d\vec{x} - \int_{\vec{a}}^{\vec{u}} \vec{F} \cdot d\vec{x}
=\int_{\vec{a}}^{\vec{u}+h\vec{v}} \vec{F} \cdot d\vec{x} + \int_{\vec{u}}^{\vec{a}} \vec{F} \cdot d\vec{x}
= \int_{\vec{u}}^{\vec{u}+h\vec{v}} \vec{F} \cdot d\vec{x} $!
\item Because of path independence, we can parametrize this as $\int_{t=0}^{t=h} \vec{F}(\vec{u} + t\vec{v}) \cdot \frac{d}{dt}[\vec{u} + t\vec{v}]dt$\
\item $= \vec{v} \cdot \int_{t=0}^{t=h} \vec{F}(\vec{u} + t\vec{v})dt$, which under $\lim \frac{1}{h}$ and L'Hopital's, becomes $\vec{v}\cdot \vec{F}(\vec{u})$.
\end{itemize}

Also: 

\fbox{$\int_C \vec{F} \cdot d\vec{x}$ depends only on the endpoints of $C$ if and only if $\int_L \vec{F} \cdot d\vec{x} = 0$ for any closed loop $L$}

\begin{itemize}
\item If F depends only on the endpoints, then any closed loop L is basically a $C_+$ and a $C_-$, so the circuit integrates to 0.
\item If every loop is 0, then any two curves with the same endpoints form a loop (after reversing one), so paths are independent. 
\end{itemize}

Note that the two theorems together imply \fbox{$\vec{F}  = \nabla f \Leftrightarrow \int_L \vec{F} \cdot d\vec{x} = 0$ for a connected domain $D$.}  Such an F is called \textbf{conservative}.

\subsection{3.5: Curl}

\textbf{Conservative} fields, where any loop line-integrates to zero, get their name because no work is performed over such a loop. 

\textbf{Main idea}: It's hard to show every loop is zero.  Instead, lets look at very small loops and go from there.  Proof idea:
\begin{itemize}
\item Pick a point $\vec{p}$ in a connected (no holes) domain D.
\item Use the right-hand rule (for example) and build a small circular loop $L$ in a plane around D.
\item Consider the linear approximation: The approximate field around $p$ is the exact field at $p$ plus the derivative at p multiplied by the delta from $p$. 
\item So, this means $\vec{F}(\vec{x}(t)) \approx \vec{F}(\vec{p}) + D\vec{F}(\vec{p})(\vec{x}(t)-\vec{p})$
\item Substitute into line integral: $\int_L [\vec{F}(\vec{p}) + D\vec{F}(\vec{p})(\vec{x}(t)-\vec{p})]dx = 0$
\item The first term is zero because it's a conservative field: $\vec{F}(\vec{p}) = \nabla (\vec{F}(\vec{p})\vec{x})$
\item The integral of the second term relates to $\frac{\delta F_y}{\delta x} -\frac{\delta F_x}{\delta y}$:
\begin{itemize}
\item \textbf{Goal}: Calculate $\frac{1}{\pi\epsilon^2} \int_L D\vec{F}(p)(\vec{x} - \vec{p})\cdot d\vec{x}$
\item Consider a loop in the xy plane only (meaning $\hat{n} = \hat{i}$). Then $\vec{x}(t) = \langle \epsilon \cos(t), \epsilon \sin(t), 0 \rangle$
\item $\vec{x}'(t) =  \langle -\epsilon \sin(t), \epsilon \cos(t), 0 \rangle$
\item $D\vec{F}(\vec{p})(\vec{x}(t)-\vec{p}) \cdot d\vec{x} 
=  \frac{\delta F_x}{\delta x}(\vec{p})(\vec{x}_x-\vec{p}_x)\frac{dx}{dt} + \frac{\delta F_y}{\delta x}(\vec{p})(\vec{x}_x-\vec{p}_x)\frac{dx}{dt}
+ \frac{\delta F_x}{\delta y}(\vec{p})(\vec{x}_y-\vec{p}_y)\frac{dy}{dt} + \frac{\delta F_y}{\delta y}(\vec{p})(\vec{x}_y-\vec{p}_x)\frac{dy}{dt}$
\item (Note that the $-p_x, -p_y$ terms will integrate to zero, and that $F_z = 0$).
\item \textbf{A-ha}: Also note that same-variable terms disappear too: $\int_L \frac{\delta F_x}{\delta x}(\vec{p})(\vec{x}_x*\frac{dx_x}{dt})  = \int_{t=0}^{t=2 \pi}\epsilon \cos(t) * -\epsilon \sin(t) = -\epsilon^2 \int_{t=0}^{t=2 \pi} \sin(t) = 0$
\item So we're left with $\int_{t=0}^{t=2\pi} \frac{\delta F_x}{\delta y} \epsilon \sin(t) (-\epsilon(\cos(t))) +  \frac{\delta F_y}{\delta x} \epsilon \cos(t) \epsilon(\cos(t))$ 
\item Using $\int_{t=0}^{t=2\pi} \cos^2(t) = \int_{t=0}^{t=2\pi} \frac{1}{2} - \frac{\cos(2t)}{2} = \pi$ (and given the curve shape, same integral result for $\sin$)...
\item $\epsilon^2(\frac{\delta F_y}{\delta x}(p) \int_{t=0}^{t=2\pi} \cos^2(t)dt - \frac{\delta F_x}{\delta y}(p)\int_{t=0}^{t=2\pi} \sin^2(t)dt )$
\item $=\pi\epsilon^2[\frac{\delta F_y}{\delta x}(p) - \frac{\delta F_x}{\delta y}(p)]$
\item \textbf{Don't Forget}: What we proved is that the line integral around that point is (a piece of) the curl times the area of the contained shape!
\end{itemize}
\item Do this again for xz and yz planes ( $\hat{n} = \hat{j}, \hat{n} = \hat{k}$), and we end up with $\frac{1}{\pi\epsilon^2}\int_L D\vec{F}(p)(\vec{x} - \vec{p})$ = (many equivalent forms)
\begin{itemize}
\item $\{-\frac{\delta F_y}{\delta z} + \frac{\delta F_z}{\delta y}, \hat{n} = \hat{i}; -\frac{\delta F_z}{\delta x} + \frac{\delta F_x}{\delta z}, \hat{n} = \hat{j}; -\frac{\delta F_x}{\delta y} + \frac{\delta F_y}{\delta x}, \hat{n} = \hat{k}\}$
\item $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
\frac{\delta}{\delta x} &\frac{\delta}{\delta y} &\frac{\delta}{\delta z}  \\
F_x & F_y & F_z \\
\end{pmatrix} \cdot \hat{n}$
\item $(\nabla \times \vec{F}) \cdot \hat{n}$
\end{itemize}
\item So the \textbf{curl} is defined as  $(\nabla \times \vec{F}) \cdot \hat{n}$, sort of like the \textbf{divergence} is $\nabla \cdot \vec{F}$
\item \emph{TODO What's the deal with the $\hat {n}$ part?}
\end{itemize}

A \textbf{potential} is a function $f$ that has a corresponding conservative vector field $\vec{F} = \nabla f$.   (Note: Wouldn't any $f$ fit the bill here?).

The big theorem for conservative fields $\vec{F}$: \fbox{$\nabla \times \vec{F} = \vec{0} \Leftrightarrow \int_L \vec{F}d\vec{x} = 0 \Leftrightarrow \vec{F} = \nabla f$   }

Example: A disk rotating around $z$ axis with rotational velocity $\omega$.  What's the curl?
\begin{itemize}
\item $\vec{r} = $ some $<x, y, z>$.
\item Velocity vector on disk $\vec{v}$ is perpendicular to disk position $r$ (perp to both $xy$ position and $z$), and is $[\omega \hat{k}] \times \vec{r}$.
\item This cross product $\vec{v} = \omega x \hat{j} - \omega y \hat{i}$.
\item So curl is $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
\frac{\delta}{\delta x} &\frac{\delta}{\delta y} &\frac{\delta}{\delta z}  \\
\omega y & \omega x& 0 \\
\end{pmatrix} = 2\omega \hat{k}$
\item \textbf{Curl Intuition}: So the curl ``curls'' around the axis of rotation, gets bigger with a faster rotation, and would reverse if the direction reversed. 
\end{itemize}

Example: A river flowing with field $\vec{F} = e^{-x^2} \hat{i}$  What's the curl?
\begin{itemize}
\item  $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
\frac{\delta}{\delta x} &\frac{\delta}{\delta y} &\frac{\delta}{\delta z}  \\
e^{-x^2} & 0 & 0 \\
\end{pmatrix} = 0$
\item  \textbf{Intuition}: If you put a cross-shaped boat in the water, axis aligned, it wouldn't twist.
\end{itemize}

Example: A river flowing with field $\vec{F} = e^{-x^2} \hat{j}$  What's the curl?
\begin{itemize}
\item  $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
\frac{\delta}{\delta x} &\frac{\delta}{\delta y} &\frac{\delta}{\delta z}  \\
0 & e^{-x^2} & 0 \\
\end{pmatrix} = -2xe^{-x^2}\hat{k}$
\item  \textbf{Intuition}: If you put a cross-shaped boat in the water, axis aligned, it WOULD twist, since the upward flows have different strengths on different sides of the boat.
\end{itemize}

\subsection{4.6: Stokes's Theorem}

Like the Divergence Theorem stacking cubes, Stokes's Theorem can be built up from small pieces too (rectangular loops). Proof:

\begin{itemize}
\item Note: ``$\delta$'' means the boundary of a region.
\item Same-oriented square loops $R_1, R_2$ can be merged to share two canceling edges (going in opposite directions) and thus one bigger loop. 
\item Therefore, $\int_{\delta [R1+R_2]} \vec{F}\cdot d\vec{x} = \int_{\delta R_1} \vec{F}\cdot d\vec{x}  + \int_{\delta R_2} \vec{F}\cdot d\vec{x} $ 
\item Before, in section 4.5, we proved that $\frac{1}{A_i} \int_{\delta r_i} \vec{F} \cdot d\vec{x} = (\nabla \times \vec{F}(p_i))\Rightarrow \int_{\delta r_i} \vec{F} \cdot d\vec{x} = A_i (\nabla \times \vec{F}(p_i))$ 
\item so $\int_{\delta [r_1 + r_2 ...]} \vec{F} \cdot d\vec{x} = \sum  A_i \nabla \times \vec{F}(p_i)$
\item As we increase the granularity and shrink $A_i$ to zero, this becomes \textbf{Stokes's Theorem}: \fbox{$\int_{\delta S} \vec{F}\cdot d\vec{x} = \iint_S \nabla \times \vec{F} \cdot \vec{dA} $}.  So the line integral of the skirt is the same as the curl integrated over the surface.
\item So this means that closed surfaces sum to zero curl, and a conservative fields sums to zero curl ($\nabla \times (\nabla F) = \nabla \times \vec{0} = \vec{0}$.
\item TODO: What happened to $\hat{n}$?
\end{itemize}

Example: The hemisphere $z = R^2- x^2 - y^2$ with equator intersecting the $xy$-plane at circle $C$, with ``arrows'' oriented counterclockwise viewed from the top of $\hat{k}$.  If $\vec{F} = -\frac{1}{3}y^3\hat{i} + \frac{1}{3}x^3\hat{j} + z\hat{k}$, compute line integral $int_C \vec{F}\cdot d\vec{x}$.  

Line integral technique:
\begin{itemize}
\item $x(t) = \langle R\cos(t), R\sin(t), 0 \rangle$, so $\vec{F} \cdot d\vec{x} = \frac{1}{3}R^4\sin^4(t) + \frac{1}{3}R^4\cos^4(t)$
\item Given identity $\int_{\theta = 0}^{2\pi}\cos^4(\theta) = \int_{\theta = 0}^{2\pi}\sin^4(\theta) = \frac{3\pi}{4}$
\item Then integral quickly becomes $ \frac{1}{3}R^4 * ( \frac{3\pi}{4} * 2) = \frac{R^4\pi}{2}$
\end{itemize}

Curl technique:
\begin{itemize}
\item $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
\frac{\delta}{\delta x} &\frac{\delta}{\delta y} &\frac{\delta}{\delta z}  \\
-\frac{1}{3}y^3& \frac{1}{3}x^3 & 0 \\
\end{pmatrix} = (x^2 + y^2)\hat{k}$
\item $f = z + x^2 + y^2 = R^2$ (level set).  Gradient $\nabla f = \langle 2x, 2y, 1\rangle$ is normal by level set and $\hat{n} = \frac{\langle 2x, 2y, 1\rangle}{\sqrt{4x^2 + 4y^2 + 1}}$, so $\hat{k} \cdot \hat{n}  = \frac{1}{\sqrt{4x^2 + 4y^2 + 1}}$
\item $dA = \|\vec{x}_x \times \vec{x}_y\| = \| \langle 1, 0, -2x \rangle \cdot \langle 0, 1, -2y \rangle \| = \sqrt{4x^2 + 4y^2 + 1}$
\item $\iint_S \nabla \times \vec{F} \cdot \vec{dA} = \iint_S (x^2 + y^2)\hat{k}\cdot\hat{n}dA = \iint_S (x^2 + y^2)(\hat{k}\cdot\hat{n})dA$
\item $=\iint_S [x^2+y^2]\frac{1}{ \sqrt{4x^2 + 4y^2 + 1}}  \sqrt{4x^2 + 4y^2 + 1}$
\item Switch to polar, don't forget Jacobian for dA: $=\int_{r=0}^R\int_{\theta = 0}^{2\pi} (r^2\cos^2(\theta)+ r^2\sin^2(\theta))r dr d\theta$
\item $ = 2\pi \int_{r=0}^R r^3 = \frac{2\pi R^4}{4} = \frac{R^4\pi}{2}$
\end{itemize}

Another feature of Stokes: If finding the surface area of a figure $S$ only depends on boundary $\delta S$, \fbox{you can switch to another figure $S'$ as long as the boundary doesn't change!  }

Example: Same unit hemisphere, but raised by a unit cylinder, with $\vec{F} = \langle e^{yz}, x\cos(z^3) + xze^{yz}, cos(sin(xyz)) \rangle$
\begin{itemize}
\item The a-ha is that a unit disk will have the same boundary, so set $z=0$
\item This reduces to $\vec{F} = 1, x, 1$, and easy to compute $\nabla \times \vec{F} = 1 * \hat{k}$
\item $\hat{n} = \hat{k}, so \iint_S \hat{n} \cdot \hat{k} = \iint_S 1 = \pi$ by area of a circle!
\end{itemize}

Another implication: a closed surface (say sphere $S$) has $ \iint_S \nabla \times \vec{F} \cdot \vec{dA}  = 0$ by the argument:
\begin{itemize}
\item Separate into two hemispheres $S_+, S_-$ with skirts running in opposite directions.
\item These have $\iint_{S_+}  \nabla \times \vec{F} \cdot \vec{dA} = \int_{C_+} \vec{F}\cdot d\vec{x} = - \int_{C_-} \vec{F}\cdot d\vec{x} = \iint_{S_-} \nabla \times \vec{F} \cdot \vec{dA}$
\item $\int_{C_+}$ and $\int_{C_-}$ therefore sum to zero.
\item And by Stokes theorem, their surface integrals $\iint_{S_+}$ and $\iint_{S_-}$ do too.
\end{itemize}

\textbf{Stokes's Theorem also works in 2D}.  Consider any such region $S$ a 2D $xy$-bound object with normal $\hat{k}$. 
\begin{itemize}
\item Write the field as $\vec{F} = \langle P(x,y), Q(x,y), 0 \rangle$
\item $\int_C \vec{F} \cdot d\vec{x} = \int_C P dx + Qdy$ by definition.
\item Take the curl, which is $\nabla \times \vec{F} = \frac{\delta Q}{\delta x} -  \frac{\Delta p}{\delta y}$
\item By Stokes's Theorem, $ \int_C P dx + Qdy = \iint_S [\frac{\delta Q}{\delta x} -  \frac{\Delta p}{\delta y}]dxdy$
\item \textbf{Green's theorem} \fbox{$\iint_D [\frac{\delta G_x}{\delta x} -  \frac{\delta G_y}{\delta y}]dx dy = \int_C \vec{G} \cdot \hat{n} ds$} is just a 2D divergence theorem.
\begin{itemize}
\item Why? Substitute $P = -G_y, Q = G_x$ into above, see $\int_C P dx + Q dy = \vec{G} \cdot \langle dy, -dx \rangle$.  The last vector is normal to $\langle dx, dy \rangle$, and unit, so we've got it. 
\end{itemize}
\end{itemize}

\textbf{Gotcha} Even though conservative flows have curl = $\vec{0}$,  $\vec{0}$ flows CAN be non-conservative on non-simple domains.  Example: 
\begin{itemize}
\item $\vec{F} = \langle \frac{-y}{x^2+y^2}, \frac{x}{x^2+y^2}, 0 \rangle $.  
\item Domain $D = \mathbb{R}^2 - (0,0)$
\item Curl $[\frac{\delta}{\delta x}(\frac{x}{x^2+y^2}) + \frac{\delta}{\delta y}(\frac{x}{x^2+y^2})]\hat{k} = \vec{0}$ after calcuation.
\item But $\int_C \vec{F}\cdot d\vec{x}$, when changed to polar with $\vec{x} = \langle \cos(t), \sin(t), 0\rangle$, becomes $\int_C \frac{\sin^2(t)}{1} + \frac{\cos^2(t)}{1^2} = \int_C 1 = 2\pi$
\item This closed loop is not 0, so the field is not conservative, but the curl still is $\vec{0}$.
\item The \textbf{gotcha} is that the domain is not simply connected like a sphere, but more like a donut with a hole in the middle.
\item If we can't shrink our circle to a point, the loop doesn't have to integrate to zero.  (TODO: Like in complex analysis?)
\end{itemize}

\subsection{4.7: Swirls and Curls}

To review:
\begin{itemize}
\item Divergence: How many lines enter or leave the neighborhood of the point, indicates how much volume changes over the flow.
\item Divergence theorem relates Flux through a closed bounding surface and the divergence over the internal volume. Main idea: stuff inside cancels out, and what's going in or out of the skin ends up being the sum of the volume's field inside.
\item Divergence theorem relates any $n-1$ and $n$ dimension hyperspaces: \fbox{ $\iint_S \vec{F} \cdot \hat{n} dA = \iiint_R \nabla \cdot \vec{F} dV$}
\item Curl: How much is the flow swirling around the point (not entering or leaving).  Rotational torque ends up motivating this in this chapter.
\item Curl relates a line integral and the 2D-surface (possibly in higher space) it skirts.
\item Stokes's Theorem: \fbox{$\int_{\delta S} \vec{F} \cdot d\vec{x} = \iint_S \nabla \times \vec{F} \cdot \vec{dA}$}\item \fbox{$$}
\item \textbf{NOTE: These cross over} in the case where field is the curl of some other field: $\vec{G} = \nabla \times \vec{F}$:
\item $ \iint_S \nabla \times \vec{F} \cdot \vec{dA} = \iint_S \vec{G} \cdot \hat{n} dA = \iiint_R \nabla \cdot (\nabla \times \vec{F}) dV = 0$
\item This is because the divergence of a curl is always zero.  Can work out in a straightforward way.  Not sure of intuition yet.
\end{itemize}

Example: Torque $\vec{\tau}$ on a mass at position $\vec{r}$ under angular velocity around origin $\vec{\omega}$ and and force $\vec{F}= \langle F_x, F_y, 0\rangle$.
\begin{itemize}
\item Torque formula: $\vec{\tau} = \vec{r} \times \vec{F}$, where $\vec{F}$ is proportional to $\frac{\delta \vec{\omega}}{dt}$
\item Rod connecting mass to origin is vector $\langle l\cos(\theta_0),  l\sin(\theta_0), 0\rangle$
\item Just compute the cross product: $\det\begin{pmatrix}
\hat{i} &\hat{j} &\hat{k}  \\
l\cos(\theta_0) &  l\sin(\theta_0) &  0 \\
F_x & F_y &  0 \\
\end{pmatrix}  = l[F_y\cos(\theta_0) -F_x\sin(\theta_0)]\hat{k}  $
\item This indicates the torque is perpendicular to $\mathbb{R}^2$ always, confining the rotation and position there.
\end{itemize}

Example: Cross-shaped paddle wheel ($\vec{r_0}$ with spokes $\vec{r_1} ... \vec{r_4}$) in a water flow $\vec{V}$
\begin{itemize}
\item Torque on paddle $\vec{\tau_i} = k[\vec{r_i} - \vec{r_0}] \times [\vec{V}(\vec{r_i}) - \vec{V}(\vec{r_0})^\perp] =  k[\vec{r_i} - \vec{r_0}] \times [\vec{V}(\vec{r_i}) - \vec{V}(\vec{r_0})]$ since the parallel part zeroes out.
\item Torque on paddle $\vec{\tau_1} = k\langle l\cos(\theta), l\sin(\theta), 0\rangle \times
\langle [\frac{\delta V_x}{\delta x} l \cos(\theta) + \frac{\delta V_x}{\delta y} l \sin(\theta)],
[\frac{\delta V_y}{\delta x} l \cos(\theta) + \frac{\delta V_y}{\delta y} l \sin(\theta)], 0\rangle$
\item $= kl^2[\frac{\delta V_y}{\delta x}\cos^2(\theta) + [\frac{\delta V_y}{\delta y} - \frac{\delta V_x}{\delta x}]\sin(\theta)\cos(\theta) - \frac{\delta V_x}{\delta y} \sin^2(\theta)]\hat{k}$
\item Recognizing that $r_2, r_3, r_4$ are the same as $r_1$ except with $\theta$ augmented by $\frac{pi}{2}, \pi, \frac{3\pi}{2}$, we eventually can sub and get 
$\vec{\tau} = \vec{\tau_1} + \vec{\tau_2} + \vec{\tau_3} +\vec{\tau_4} = 2kl^2[\frac{\delta V_y}{\delta x} - \frac{\delta V_x}{\delta y}]\hat{k}$
\item So, $\vec{\tau}$ is proportional to $[\frac{\delta V_y}{\delta x} - \frac{\delta V_x}{\delta y}]\hat{k}$
\item Fact, we define \textbf{curl in 2D} as \fbox{$curl(\vec{V}(x,y) = [\frac{\delta V_y}{\delta x} - \frac{\delta V_x}{\delta y}]\hat{k}$}
\item This can just as easily be in the yz-plane as $curl(\vec{V}(y,z) = [\frac{\delta V_z}{\delta y} - \frac{\delta V_y}{\delta z}]\hat{i}$
\item Note: we also define a \textbf{right-handed} system as one where vectors $\vec{v_1} \times \vec{v_2} = \vec{v_3}$, so the $j$ one ends up being negative since $\hat{i} \times \hat{k} = -\hat{j}$
\end{itemize}

An \textbf{irrotational} field is one where the cross-shaped paddles don't spin, or where curl = $\vec{0}$.

\subsection{4.8: Differential Forms}

\textbf{General Form of Stokes' Theorem} is \fbox{$\int_M d\omega = \int_{\delta M}\omega$}.  This chapter explains it.

A \textbf{1-form} takes a \emph{vector}, outputs a \emph{number} and is \emph{linear}.  So \fbox{$\omega: \mathbb{R}^n\rightarrow \mathbb{R}, \omega[a\vec{x} + b\vec{y}] = a\omega[\vec{x}]  + b \omega [\vec{y}].$}
\begin{itemize}
\item 1-form: $\vec{v} \cdot \vec{x}$, since $\vec{v} \cdot (a\vec{x} + b\vec{y}) = a\vec{v} \cdot \vec{x} + b\vec{v} \cdot \vec{y}$
\item Not a 1-form: $\|x\|^2$, since $\sqrt{((ax_1)^2 + (by_1))^2) + (ax_2)^2 + (by_2))^2) + ... }^2
 = a^2(x_1^2 + x_2^2) + b^2(y_1^2 + y_2^2) + 2ax_1by_1 + 2ax_2by_2 \neq a^2(x_1^2 + x_2^2) + b^2(y_1^2 + y_2^2) 
 = \sqrt{ax_1^2 + ax_2^2}^2 + \sqrt{by_1^2 + by_2^2}^2$. Extra crossover term.
\item Remember $proj_v(\vec{x}) = \frac{\vec{x} \cdot \vec{v}}{\vec{v} \cdot \vec{v}}\vec{v}$
\item So 1-form: $proj_v(x) \cdot \vec{u}$, because it's just $\vec{x} \cdot$ a bunch of stuff.
\item However, GOTCHA, not a 1-form: $proj_v(x)$, since it outputs a vector, not a number!
\item Obviously $proj_x(\vec{v}) = \frac{\vec{v} \cdot \vec{x}}{\vec{x} \cdot \vec{x}}\vec{x}$ is not, as both numerator and denominator are quadratic.
\item An index into a vector is a one form: $[0, 1, 0] \cdot [x,y z ] =y$
\item A mean is a one form: $[\frac{1}{n}, \frac{1}{n}, \frac{1}{n}] \cdot  [x,y,z ]$
\item Note: A zero-form is a function on $\vec{x}$, since it just returns a scalar at that point.
\end{itemize}

Define $dx_j[\hat{e}_i] = 1$ if $i=j$ else $0$.  Note: This is really a function $dx_j(\vec{x}) = [0, 0, 0, ... 1, ... 0] \cdot \vec{x}$, where the 1 is in the $j$ spot. These are the \textbf{basis 1-forms} because 
\\
\fbox{any 1-form can be written as $\omega = a_1dx_1 + ... + a_ndx_n$.}.  
So $\omega[\vec{x}] = \vec{v} \cdot \vec {x}$ is just $\sum_{j=1}^n (v_jdx_j)[\vec{x}].$

\textbf{NOTE}: This 1-form $dx_2$, say, looks an awful lot like $dy$, since $\vec{v} = <x, y, z> = <0, y, 0> \rightarrow dy[\vec{v}] = 1$

A \textbf{tensor} is a linear function to a number taking not just one but multiple vectors as inputs.  
\\ 
The \textbf{tensor product} is built from linear combinations of $(dx_i \otimes dx_j)(\vec{v},\vec{w}) = dx_i(\vec{v}) \cdot dx_j(\vec{w}) = v_iw_j$, so $T(\vec{v}, \vec{w}) = (\sum_{i,j} t_{ij}dx_i \otimes dx_j)(\vec{v}, \vec{w}) = \sum_{i.j}t_{ij}v_iw_j$.  Just combos of each pair.  Examples and non-examples of 2-tensors:

\begin{itemize}
\item Tensor: $\vec{x} \cdot \vec{y}$.  Linear, outputs a number.
\item Non-Tensor: $\vec{x} + \vec{y}$.  Linear, but outputs a vector.
\item Non-Tensor: $\vec{x} \times \vec{y}$.  Linear, but outputs a vector.
\item Non-Tensor: $\|\vec{x}\| + \| \vec{y}\|$. Non-linear
\end{itemize}

As for 3-tensors:
\begin{itemize}
\item $\det(\vec{u}, \vec{v}, \vec{w})$ defined as $\det\begin{pmatrix}
u_1 & v_1 & w_1 \\
u_1 & v_2 & w_2 \\
u_1 & v_3 & w_3
\end{pmatrix}$ is linear in changes to $u, v, w$.
\item $(\vec{u} \times \vec{v})\cdot \vec{w}$ actually IS  $\det\begin{pmatrix}
w_1 & w_2 & w_3 \\
u_1 & u_2 & u_3 \\
v_1 & v_2 & v_3 \\
\end{pmatrix}$ 
\item $(\vec{u} \times \vec{v}) \times \vec{w}$ outputs a vector.
\item $(\vec{u} \cdot \vec{v})\vec{w}$ outputs a vector.
\item A 3-tensor is generally $T(\vec{u}, \vec{v}, \vec{w}) = (\sum_{i,j,k} t_{ijk}dx_i \otimes dx_j \otimes dx_k)(\vec{i}, \vec{v}, \vec{w}) = \sum_{i.j,k}t_{ijm}u_iv_jw_k$, so think of it as $(dx_i \otimes dx_j \otimes dx_k)(\vec{u}, \vec{v}, \vec{w}) = u_iv_jw_k$
\end{itemize}

Side note: Wedge product $\vec{a} \wedge \vec{b} =\frac{1}{2}[ \vec{a} \otimes \vec{b} - \vec{b} \otimes \vec{a}]$ is a square matrix measure of the anti-commutativity of $\vec{a} \otimes \vec{b}$.  So, $(\vec{a} \wedge \vec{b})_{ij} = u_iv_j - u_jv_i$. (Note: Some texts seem to omit the $\frac{1}{2}$.) It has algebraically verifiable properties of:

\begin{itemize}
\item Antisymmetry: $(\vec{a} \wedge \vec{b}) = -(\vec{b} \wedge \vec{a})$.  
\begin{itemize}
\item This makes the tensor called \textbf{alternating}. 
\item This also means $(\vec{a} \wedge \vec{a}) = -(\vec{a} \wedge \vec{a}) = [0]$
\item This also means (with basic associativity) that switching any two flips the sign: $\vec{a} \wedge \vec{b} \wedge \vec{c} = -\vec{b} \wedge \vec{a} \wedge \vec{a}$
\end{itemize}
\item Bilinearity:  $(c\vec{a} \wedge \vec{b}) = c(\vec{a} \wedge \vec{b})$
\item Distributivity: $\vec{a} \wedge (\vec{b} + \vec{c} ) = \vec{a} \wedge \vec{b} + \vec{a} \wedge \vec{c}   $
\item And others.
\end{itemize}

A \textbf{2-form} is an alternating 2-tensor: a linear operation taking in 2 vectors and producing a number, where switching the vectors flips the sign.

A \textbf{3-form} is an alternating 3-tensor: a linear operation taking in 3 vectors and producing a number, where flipping two vectors flips the sign, so ``permutation rules'' signage.  So switch 

A three dimensional wedge product continues this and again has all combinations :

$dx_i \wedge dx_j \wedge dx_k = \frac{1}{6}[ \\
	dx_i \otimes dx_j \otimes dx_k - dx_i \otimes dx_k \otimes dx_j \\
	+ dx_j \otimes dx_k \otimes dx_i - dx_j \otimes dx_i \otimes dx_k \\
	+ dx_k \otimes dx_i \otimes dx_j - dx_k \otimes dx_j \otimes dx_i ]$
	
\textbf{Form fields} eat position $\vec{x}$ and return a form (tensor taking in vectors and returning a number).
\begin{itemize}
\item Main rule: $df = \sum_{j=1}^n \frac{\delta f}{\delta x_j} dx_j$.
\item Implies somehow (?): $d( \sum_{j=1}^n f_j(\vec{x})dx_j) = \sum_{j=1}^n [df_j] \wedge dx_j$.
\item And suspending disbelief this means $d( \sum_{i,j=1}^n f_{ij}(\vec{x})dx_i \wedge dx_j ) = \sum_{i,j=1}^n [df_{ij}] \wedge dx_i \wedge dx_j$.
\item Example: If $\omega = x dy - y dx = -y dx + x dy$
\begin{itemize}
\item Set $f_1  = -y, f_2 = x$
\item $d\omega = df_1 \wedge dx + df_2 \wedge dy$ by second rule
\item $=(\frac{\delta f_1}{\delta x} dx + \frac{\delta f_1}{\delta y} dy) \wedge dx + (\frac{\delta f_2}{\delta x} dx + \frac{\delta f_2}{\delta y} dy) \wedge dy$ by main rule
\item $-dy \wedge dx + dx \wedge dy = 2dx \wedge dy$ by calcuation
\end{itemize}
\end{itemize}

Another example: $\omega = \frac{1}{2}(x^2+y^2+z^2)[dx\wedge dy + dx\wedge dz + dy\wedge dz]$ sees $d\omega$ as:
\begin{itemize}
\item $=[x dx + y dy z dz] [dx\wedge dy + dx\wedge dz + dy\wedge dz]$ by third rule
\item $=x (dx \wedge dy \wedge dz) + y (dy \wedge dx \wedge dz )+ z (dz \wedge dx \wedge dy)$ since anything like $dx \wedge dx$ zeroes out
\item $=(x - y + z) dx \wedge dy \wedge dz $ 
\end{itemize}

\textbf{Manifolds} are part of differential geometry, or really, just space curves (1-manifold), surfaces (2-manifolds), regions (3-manifolds).  The number tells us how many numbers are needed to specify a point within it.

Rule: We can only integrate an n-form on an n-manifold.  So, this means we can integrate a single (linear) vector-eating function on a line, a tensor eating two vectors on a surface, a 3-form on a region.

Be careful - it's not just the variable count: $\omega = x dx - y dz \rightarrow d\omega = (dx) \wedge dx - (dy) \wedge dz = -dy \wedge dz$.  So it is a 2-manifold, and we can only integrate $d\omega$ on a surface.

Example: If we want to integrate $f(x,y,z) = x^2 + 2y^2 + 3z$ on $\vec{x}(t) = \langle 1-t, t, 2+3t\rangle t \in [0,1]$:

\begin{itemize}
\item $dx = \frac{dx}{dt}dt, dx = \frac{dy}{dt}dt, dz = \frac{dz}{dt}dt$
\item $\int_M df = \int_M[2xdx + 4ydy + 6zdz]$
\item $= \int_{t=0}^1 [2x(t) \frac{dx}{dt} + 4y(t)\frac{dy}{dt} + 6z(t) \frac{dz}{dt}]$
\item $= \int_{t=0}^1 [2(1-t)(-1) + 4(t)(1) + 6(2+3t)(3)]dt = 64$
\end{itemize}

Example: $M = x^2 + y^2 + z^2 \leq 1$, $\omega = x dy \wedge dz -ydx\wedge dz + z dx \wedge dy$.  What's $\int_{\delta M}\omega$

\begin{itemize}
\item Parameterize
\item $dx = \frac{\delta}{\delta u}(\sin(u)\cos(v)) + \frac{\delta}{\delta v}(\sin(u)\cos(v)) = \cos(u)\cos(v)du-\sin(u)\sin(v)dv$
\item $dy = \frac{\delta}{\delta u}(\sin(u)\sin(v)) + \frac{\delta}{\delta v}(\sin(u)\sin(v)) = \cos(u)\sin(v)du+\sin(u)\cos(v)dv$
\item $dz = \frac{\delta}{\delta u}(\cos(u)) = -\sin(u)du$
\item $dx \wedge dy = ( \cos(u)\cos(v)\sin(u)\cos(v) + \sin(u)\sin(v) \cos(u)\sin(v)) du \wedge dv =  \cos(u)\sin(u) du \wedge dv $
\item $dx \wedge dz = -\sin(u)\sin(v)\sin(u) du \wedge dv =  -\sin^2(u)\sin(v) du \wedge dv $
\item $dy \wedge dz = -\sin(u)\sin(v)\sin(u) du \wedge dv =  \sin^2(u)\cos(v) du \wedge dv $ (wrong should be negative)?
\item $\omega = x dy \wedge dz -ydx\wedge dz + z dx \wedge dy$.
\item  $= \sin(u)\cos(v) \sin^2(u)\cos(v) du \wedge dv - \sin(u)\sin(v)[  -\sin^2(u)\sin(v) du \wedge dv] + \cos(u) \cos(u)\sin(u) du \wedge dv$
\item $=[\sin^3(u) + \cos^2(u)\sin(u)]  du \wedge dv$
\item $= \sin(u)  du \wedge dv$
\item So $\int_{v=0}^{2\pi}\int_{u=0}^{u=\pi} \sin(u)dudv  = 4\pi$
\end{itemize}

Example: Do it again by computing the other way $\int_M \delta\omega$:

\begin{itemize}
\item $\omega = x dy \wedge dz -ydx\wedge dz + z dx \wedge dy$.  
\item $\delta \omega = 3 dx \wedge dy \wedge dz$
\item $\iint_M 3dxdydz = 3 \times \frac{4\pi}{3} = 4\pi$.
\item \textbf{BIG}: This is one example of $\int_{\delta M} \omega = \int_M d\omega$
\end{itemize}

Note: The great unification happens when dumping different forms into $\int_{\delta M} = \int_M \delta \omega$

\begin{itemize}
\item If we use $\omega = F_x dy \wedge dz + F_y dz \wedge dx + F_z dx \wedge dy$, then we find that $d\omega = \nabla \cdot F dx \wedge dy \wedge dz$.
\item Then $\int d\omega = \iiint_M  \nabla \cdot F dx dy  dz$. (Why drop the wedges, exactly?)
\item If we parametrize $\delta M $ as $\vec{x}(u,v)$, with $dx = \frac{\delta x}{\delta u}du + \frac{\delta x}{\delta u}dv$, similar for the other two, and compute all the $dy \wedge dz$ etc....
\item And we churn through noticing cross products inside, we get $\omega = \vec{F}(\vec{x}(u,v)) \cdot (\vec{u}_u \times \vec{x}_v)du \wedge dv$
\item So $\int_{\delta M} \omega = \iint_D \vec{F}(\vec{x}(u,v)) \cdot (\vec{u}_u \times \vec{x}_v)du \wedge dv$
\item $=\vec{F}(\vec{x}(u,v)) \cdot \frac{\vec{u}_u \times \vec{x}_v}{\| (\vec{u}_u \times \vec{x}_v)\|}  \|(\vec{x}_u \times \vec{x}_v)\|  du \wedge dv$ 
\item $=\iint_{\delta M} \vec{F} \cdot \vec{dA}$
\item So we've equated $\iiint_M \nabla \cdot F dx dy dz =\iint_{\delta M} \vec{F} \cdot \vec{dA}$ after dropping the wedges somehow.
\item Divergence Theorem!
\end{itemize}

Similarly, if we use $\omega = F_x dx + f_y dy + F_z dz$, we can do something similar to get $\int_{\delta M} \vec{F} \cdot d\vec{x} = \iint_M \nabla \times \vec{F} \cdot \vec{dA}$, or Stokes' Theorem.

\textbf{NOTE}: Still need some intuiition for these forms.  Apparently these are described as ``that which makes Stokes's and Divergence Theorem fall out'': https://math.stackexchange.com/questions/2858098/what-is-a-differential-form

Note: From lecture at ``https://www.youtube.com/watch?v=wlo2V8H5khM'', $d_{i_1, i_2, ... i_k}$ is a function that takes $k$ vectors in $\mathbb{R}^n$, concatenates horizontally, selects only rows $i_1, i_2, ... i_k$ and takes the determinant.  This is a multilinear, alternating function.

\section{Chapter 5: Applications}
\subsection{5.1: The Laplacian}

\textbf{Laplacian}: $\nabla^2 \vec{F} = \nabla \cdot \nabla \vec{F} = \frac{\delta^2}{\delta x^2}F_x + \frac{\delta^2}{\delta y^2}F_y + \frac{\delta^2}{\delta z^2}F_z$ shows up often in partial differential equations.  

 Example: Finding Potential $V$ of an electrostatic field $\vec{E} = -\nabla V$, if $\vec{E} = \frac{Q}{4\pi\epsilon_0}\frac{\vec{x}}{\|\vec{x}\|^3}$
\begin{itemize}
\item Consider this in spherical coordinates and note that V is only a function of $\rho$.
\item Use the gradient in spherical coordinates: $\nabla \cdot \vec{V} = \frac{\delta V}{\delta\rho} \hat{\rho} + \frac{1}{\rho}\frac{\delta V}{\delta \phi} \hat{\phi} + \frac{1}{\rho \sin(\phi)} \frac{\delta V_{\theta}}{\delta \theta}\hat{\theta}$, ignoring the second and third terms of the sum.
\item  $\vec{E} = \frac{Q}{4\pi\epsilon_0}\frac{\vec{x}}{\|\vec{x}\|^3} =  \frac{Q}{4\pi\epsilon_0}\frac{1}{\rho^2}\hat{\rho} = -\nabla V = \frac{\delta V}{\delta\rho} \hat{\rho}$
\item $\frac{Q}{4\pi\epsilon_0}\frac{1}{\rho^2} = -\frac{\delta V}{\delta \rho}$
\item $\frac{Q}{4\pi\epsilon_0}\frac{1}{p} + C = V(\rho)$ or $\frac{Q}{4\pi\epsilon_0}\frac{1}{\|\vec{x}\|} = V(\vec{x})$ is potential for point charge at origin.
\item This means a set of charges at points $P_i$ have total potential $\sum_i \frac{Q}{4\pi\epsilon_0}\frac{1}{\|\vec{x} - \vec{P_i}\||}$, creating the field $\vec{E} = -\nabla V$
\item So if $\nabla \cdot \vec{E} = \frac{\rho}{\epsilon_0}$ (Gauss's law), and $\vec{E} -\nabla V$, then:
\item $\nabla \cdot \vec{E} = -\nabla \cdot (\nabla V) = -\frac{\delta}{\delta x}\frac{\delta V}{\delta x} - \frac{\delta}{\delta y}\frac{\delta V}{\delta y} -\frac{\delta}{\delta z}\frac{\delta V}{\delta z} =$ \fbox{$= -\delta_x^2V -\delta_y^2V -\delta_z^2 = -\nabla^2 V$}
\item   Laplacian shorthand : $\nabla^2 = \nabla \cdot \nabla = \frac{\delta^2}{\delta x^2} +  \frac{\delta^2}{\delta y^2} +  \frac{\delta^2}{\delta z^2}$
\end{itemize}

Example: Two infinite thin sheets of metal aligned with $xy$-plane.  Function $T(z)$ is temperature, $T(0) = T_0, T(h)= 2T_0$.  What is $T$?

\begin{itemize}
\item From Laplace, $\nabla^2 T = (\frac{\delta^2}{\delta x^2} +  \frac{\delta^2}{\delta y^2} +  \frac{\delta^2}{\delta z^2})T = \frac{\delta^2T}{\delta z^2}$ since only $z$ matters here.
\item Anti-derive $T''(z) = 0$ twice to get $T(z) = Az+B$.  
\item $T(0) = T_0 \Rightarrow B = T_0 \Rightarrow T(z) = Az+T_0$
\item $T(h) = 2T_0 \Rightarrow Ah = T_0 \Rightarrow A = \frac{T_0}{h} \Rightarrow T(z) = \frac{T_0}{h}z+ T_0$
\end{itemize}

Exmple: Thin circlular disk of radius $r$ floats on a cushion of air, over centered opening of radius $\epsilon$.  Pressure at opening is $p_{in}$, at edge (and outside) is necessarily $p_{atm}$.  If also follows $\nabla^2 p(r) = 0$
\begin{itemize}
\item To get "nabla squared" (or cylindrical Laplacian equation) we want to get "nabla of a vector" (div) on "nabla of a function" (grad).
\item (1) Cylindrical grad equation: $\nabla f = \frac{\delta f}{\delta r} \hat{r} +   \frac{1}{r} \frac{\delta f}{\delta \theta} \hat{\theta} +  \frac{\delta f}{\delta z} \hat{z}$
\item (2) Cylindrical div equation: $\nabla \cdot \vec{F} = \frac{1}{r}\frac{\delta rF_r}{\delta r} \hat{r} +  \frac{1}{r} \frac{\delta F_\theta}{\delta \theta} \hat{\theta} +  \frac{\delta F_z}{\delta z} \hat{z}$
\item Subbing (1) in to (2) yields  \fbox{$\nabla \cdot (\nabla\vec{F}) = \frac{1}{r}\frac{\delta (r\frac{\delta f}{\delta r})}{\delta r} \hat{r} +  \frac{1}{r} \frac{\delta}{\delta \theta}(\frac{1}{r}\frac{\delta f}{\delta \theta}) \hat{\theta} +  \frac{\delta ^2F f}{\delta z^2} \hat{z}$} $= 0$
\item We know we don't depend on $\theta$ or $z$, so this is just $0=\frac{1}{r}\frac{\delta (r\frac{\delta f}{\delta r})}{\delta r}$
\item We know $r$ is not 0, so multiply by $r$: $0 = \frac{\delta}{\delta r} (r\frac{\delta f}{\delta r}) \Rightarrow A = r\frac{\delta f}{\delta r}  \Rightarrow \frac{A}{r}=\frac{\delta f}{\delta r} $
\item Anti-derivatives yield $A\ln(r) + B = f(r)$, and specifically
\\
$A\ln(R) +B = p_{atm}$ 
\\
$Aln(\epsilon) +B = p_{in}$
\item Subtract to get $A\ln (\frac{R}{\epsilon}) =  p_{atm}-p_{in} = \Delta p \Rightarrow A = \frac{\Delta p}{\ln (\frac{R}{\epsilon})} $
\item Sub our new $A$ in to get $B = p_{in} - \ln(\epsilon) \frac{\Delta p}{\ln (\frac{R}{\epsilon})}$
\item Subbing all gives us $p(r) = \Delta p \frac
  {\ln(\frac{r}{\epsilon})}
  {\ln(\frac{\epsilon}{R})}+ p_{in}$
\end{itemize}

Challenge: If lift $L = \iint_{disk} [p-p_{atm}]dA = W$ weight causes the disk to float, what $\Delta p = p_{in} - p_{atm}$ satisfies it?
\begin{itemize}
\item We know form $r \in [\epsilon, R], p(r) = \Delta p \frac
  {\ln(\frac{r}{\epsilon})}
  {\ln(\frac{\epsilon}{R})}+ p_{in}$ 
\item And $r \in [0, \epsilon]$ means it is $p_in$.
\item Split the integral apart. 
\item $L = \int_{\theta =  0}^{2\pi} \int_{r=0}^{r=R} p-p_{atm}dA$, or $L = 2\pi\int_{r=0}^{r=R} p(r)r dr  - \pi*R^2p_{atm}$ (term 1)
\item $2\pi\int_{r=0}^{r=R} p(r)r dr = 2\pi(\int_{r=\epsilon}^{r=R} p(r)r dr) + 2\pi(\int_{r=0}^{r=\epsilon} {p_{in}}r dr)$. Second term is $\pi * \epsilon^2 p_{in}$
\item $2\pi(\int_{r=\epsilon}^{r=R} \Delta p \frac
  {\ln(\frac{r}{\epsilon})}
  {\ln(\frac{\epsilon}{R})}) = \frac{2\pi \Delta p}{\ln(\frac{\epsilon}{R})} \int_{r=\epsilon}^{r=R} \ln(\frac{r}{\epsilon})r dr
= \frac{2\pi \Delta p}{\ln(\frac{\epsilon}{R})} \int_{r=\epsilon}^{r=R} \ln(r)r dr -  \frac{2\pi \Delta p \ln(\epsilon)}{\ln(\frac{\epsilon}{R})} *\frac{R^2-\epsilon^2}{2}$
\item Last term above is term 3.
\item Term 4 is $- \frac{R^2 - \epsilon^2}{4} + \frac{R^2\ln(R) - \epsilon^2\ln(\epsilon)}{2}$
\item $L = \frac{\Delta P}{\ln(\frac{R}{\epsilon})} \frac{\pi}{2}(R^2 -\epsilon^2) = W$
\item $\Rightarrow \Delta p = \frac{2W\ln(\frac{R}{\epsilon})}{\pi (R^2-\epsilon^2)}$
\end{itemize}

Example: Capacitor for storing energy - Two concentric conducting shells, $R_1 < R_2$ voltage $V(R_1) = 0, V(R_2) = V_0$.  No charge between the spheres so $\nabla^2V = 0$.
\begin{itemize}
\item Use spherical coordinates.  Everything obviously depends on $\rho$.  
\item (1) Spherical grad equation: $\nabla f = \frac{\delta f}{\delta \rho} \hat{\rho} +   \frac{1}{\rho} \frac{\delta f}{\delta \phi} \hat{\phi} +  \frac{1}{\rho\sin(\phi))}\frac{\delta f}{\delta \theta} \hat{\theta}$
\item (2) Spherical div equation: $\nabla \cdot \vec{F} = \frac{1}{\rho^2}\frac{\delta (\rho^2F_{\rho})}{\delta \rho}  +   \frac{1}{\rho \sin(\phi)} \frac{\delta}{\delta \phi} (\sin(\phi) F_{\phi}) +  \frac{1}{\rho\sin(\phi)}\frac{\delta F_{\theta}}{\delta \theta} $
\item Subbing (1) in to (2) yields $\nabla^2V = \frac{1}{\rho^2}\frac{\delta}{\delta \rho}[\rho^2\frac{\delta V}{\delta \rho}] + \frac{1}{\rho^2\sin(\theta)}  \frac{\delta}{\delta \phi} (\sin(\phi)\frac{\delta V}{\delta \phi})  + \frac{1}{\rho^2\sin^2(\phi)}\frac{\delta ^2V}{\delta \theta^2}$
\item Solving for the potential $V(\rho), R_1 \leq \rho \leq R_2$ is the same as above:
\item $\frac{1}{\rho^2}\frac{\delta}{\delta \rho}[\rho^2\frac{\delta V}{\delta \rho}] = 0$.  Multiply by $\rho^2$
\item $A = \rho^2 \frac{\delta V}{\delta \rho} \rightarrow A\rho^{-2}= \frac{\delta V}{\delta \rho} \Rightarrow B - A\rho^{-1} = V(\rho)$
\item Churning through you get $B = \frac{A}{R_1}, A = \frac{V_0R_1R_2} {R_2-R_1}$, and $V(\rho) = \frac{R_2V_0}{R_2-R_1}[1 - \frac{R_1}{\rho}]$
\item And therefore the field $\vec{E} = - \nabla V$ (grad of potential) and Spherical grad equation (1) above mean that $\vec{E} = \frac{\delta V}{\delta \rho}\hat{\rho} = \frac{-R_1r_2V_0}{R_2-R_1}\frac{\hat{\rho}}{\rho^2}$
\item Note that this looks like the field from a point charge (proportional to $\frac{\vec{x}}{\|\vec{x}\|^3}$) for the same 
\item Given that spherical charge distribution vs. placed all at center is the same field (from previous chapters), this applies as well since they're spherical shells. 
\end{itemize}

\subsection{8.2:Gaussian Integrals I}

Summary: A set of tricks for evaluating $I(a) = \int_{-\infty}^{\infty} x^n e^{-\frac{a}{2} x^2} dx$, specifically when $n=0: (\sqrt{\frac{2\pi}{a}})$, even (recursive built on $n=0$) or or odd (zero).

\textbf{Gaussian for n=0}:
\begin{itemize}
\item $I(a)$ is defined as a function of $a$ with dummy variable $x$( or whatever):  $I(a)= \int_{-\infty}^{\infty} e^{-\frac{a}{2} x^2} dx$
\item Trick to getting this integral is \emph{integrating it in both directions} over the $xy$ plane and taking the square root.
\item Specifically, $I(a)^2 = (\int_{-\infty}^{\infty} e^{-\frac{a}{2} x^2}dx)(\int_{-\infty}^{\infty} e^{-\frac{a}{2} y^2}dy) = (\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} e^{-\frac{a}{2} (x^2+y^2)}dx dy$
\item Rewrite in polar and solve: $I(a)^2 = \int_{\theta = 0}^{2\pi}\int_{r=0}^{\infty} e^{\frac{a}{2}r^2}rdrd\theta = [2\pi * \frac{-1}{a}e^{-\frac{a}{2}r^2}]_0^{\infty} = \frac{2\pi}{a}$
\item Thus $I(a)= \sqrt{ \frac{2\pi}{a}}$
\end{itemize}

\textbf{Gaussian for n=odd}: $\int_{-\infty}^{\infty}x^ne^{-\frac{a}{2}x^2} dx, a >0$: The $e^{f(x)}$ factor is symmetric, $x^n$ is odd, so the whole thing is zero.

\textbf{Gaussian for n=2}: 
\begin{itemize}
\item Use a derivative trick to \emph{reproduce our tricky integral} from known identity $I(a) = \sqrt{\frac{2\pi}{a}}$ instead of trying to evaluate $\int_{-\infty}^{\infty} x^2 e^{-\frac{a}{2} x^2}dx)$
\item $\frac{\delta}{\delta a} I(a) = \frac{\delta}{\delta a} \sqrt{2\pi}a^{-\frac{1}{2}} = -\frac{1}{2}\sqrt{2\pi}a^{-\frac{3}{2}}$
\item $\frac{\delta}{\delta a} I(a) = \int_{-\infty}^{\infty} \frac{\delta}{\delta a} [e^{-\frac{a}{2}x^2}] dx = -\frac{1}{2} \int_{-\infty}^{\infty} x^2 [e^{-\frac{a}{2}x^2}] dx$
\item Equating the ends of the last two lines shows that $\int_{-\infty}^{\infty} x^2 [e^{-\frac{a}{2}x^2}] dx = \sqrt{2\pi}a^{-\frac{3}{2}}$
\item From here, we can attack $n=4$, then from there, $n=6$, etc.
\end{itemize}

If we're going to evaluate the general form $\int_{-\infty}^{\infty} [e^{-\frac{a}{2}x^2-bx+c}] dx$:
\begin{itemize}
\item First, complete the square of $f = -\frac{a}{2}x^2-bx+c$
\begin{align}
-\frac{2}{a}f =x^2+\frac{2b}{a}x-\frac{2c}{a} \\
-\frac{2}{a}f =(x+\frac{b}{a})^2-\frac{b^2}{a^2}-\frac{2c}{a} \\
f = -\frac{a}{2}(x+\frac{b}{a})^2 +\frac{b^2}{2a}+c\\
\end{align}
\end{itemize}

Therefore, $\int_{-\infty}^{\infty} [e^{-\frac{a}{2}x^2-bx+c}] dx =  \int_{-\infty}^{\infty} [e^{-\frac{a}{2}(x+\frac{b}{a})^2 +\frac{b^2}{2a}+c}]dx = [e^{\frac{b^2}{2a}+c}] \int_{-\infty}^{\infty} [e^{-\frac{a}{2}(x+\frac{b}{a})^2}] dx$.  Notice that $u = x+\frac{b}{a}, du = dx$ makes the integral $\sqrt{\frac{2\pi}{a}}$.

So general form of Gaussian, with $n=0$, is \fbox{ $\int_{-\infty}^{\infty} [e^{-\frac{a}{2}x^2-bx+c}] dx =\sqrt{\frac{2\pi}{a}} [e^{\frac{b^2}{2a}+c}] $}.

\end{document}

